---
title: "Cluster-Based Relative Value Strategy: Intra-Cluster Mean Reversion on SPDR Sector ETFs"
author: "Mathew Frame"
date: "07/04/2026"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
---

```{r setup}
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(tidyquant)
library(gt)
library(cluster)
library(dlm)
library(ggrepel)
library(scales)
library(PerformanceAnalytics)
library(factoextra)
library(ggdendro)
library(patchwork)
library(plotly)
library(foreach)
library(doParallel)
library(RTL)
```

```{r data-pull}
#| message: false
#| warning: false
#| echo: false

tickers <- c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")

raw <- tidyquant::tq_get(tickers, from = "2000-06-01", to = "2026-03-31")

# Log returns grouped by symbol
returns <- raw %>%
  dplyr::group_by(symbol) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(ret = log(adjusted / dplyr::lag(adjusted))) %>%
  dplyr::ungroup() %>%
  dplyr::filter(!is.na(ret))

# Training / testing split -- date filter only, never slice()
TRAIN_END  <- "2024-12-31"
TEST_START <- "2025-01-01"

ret_train <- returns %>% dplyr::filter(date <= TRAIN_END)
ret_test  <- returns %>% dplyr::filter(date >= TEST_START)

# Sector ETFs only (SPY is benchmark, not traded)
sector_tickers <- tickers[tickers != "SPY"]

sector_train <- ret_train %>% dplyr::filter(symbol %in% sector_tickers)
sector_test  <- ret_test  %>% dplyr::filter(symbol %in% sector_tickers)
spy_train    <- ret_train %>% dplyr::filter(symbol == "SPY")
spy_test     <- ret_test  %>% dplyr::filter(symbol == "SPY")
```

------------------------------------------------------------------------

## Executive Summary

*\[To be written last. 3 sentences: strategy thesis, training result, testing result.\]*

------------------------------------------------------------------------

## 1. Intuitive Rationale

*\[Strategy thesis, clustering rationale, failure conditions, paper citations.\]*

------------------------------------------------------------------------

## 2. Data and Features

*\[Data sources table, log return justification, rolling correlation matrix as clustering input.\]*

------------------------------------------------------------------------

## 3. Clustering

*\[Elbow method to select k, hierarchical clustering on distance matrix, silhouette validation, cluster visualization, commercial interpretation.\]*

------------------------------------------------------------------------

## 4. Kalman Filter

*\[Local level model, smoothed price and slope, hard gate rule, visualization, gate firing frequency.\]*

------------------------------------------------------------------------

## 5. Signal Generation

*\[Within-cluster z-scores, inverse rank weighting, cross-cluster equal weight, worked example.\]*

------------------------------------------------------------------------

## 6. Execution and Implementation

*\[Mermaid mental model, execution rules, signal lag, transaction costs, position sizing.\]*

------------------------------------------------------------------------

## 7. Training Period Backtest

*\[Equity curve vs SPY, drawdown chart, performance table.\]*

------------------------------------------------------------------------

## 8. Risk Appetite

*\[Max drawdown tolerance, optimization objective, defensive trigger.\]*

------------------------------------------------------------------------

## 9. Optimization

*\[Grid search over parameters, heat map, selected parameters.\]*

------------------------------------------------------------------------

## 10. Testing Period Results

*\[Out-of-sample backtest, honest assessment.\]*

------------------------------------------------------------------------

## 11. Lessons Learned

*\[Min 5 points x 3-5 sentences, FIN 440 QTP failures referenced.\]*
