---
title: "Cluster-Based Relative Value Strategy: Intra-Cluster Mean Reversion on SPDR Sector ETFs"
author: "Mathew Frame"
date: "2026-04-07"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show Code"
    code-copy: true
---

```{r setup}
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(tidyquant)
library(gt)
library(cluster)
library(dlm)
library(ggrepel)
library(scales)
library(PerformanceAnalytics)
library(factoextra)
library(ggdendro)
library(patchwork)
library(plotly)
library(foreach)
library(doParallel)
library(RTL)
```

```{r data-pull}
#| message: false
#| warning: false
#| echo: false
#| cache: true

tickers <- c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")

raw <- tidyquant::tq_get(tickers, from = "2000-06-01", to = "2026-03-31")

# Log returns grouped by symbol
returns <- raw %>%
  dplyr::group_by(symbol) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(ret = log(adjusted / dplyr::lag(adjusted))) %>%
  dplyr::ungroup() %>%
  dplyr::filter(!is.na(ret))

# Training / testing split -- date filter only, never slice()
TRAIN_END  <- "2024-12-31"
TEST_START <- "2025-01-01"

ret_train <- returns %>% dplyr::filter(date <= TRAIN_END)
ret_test  <- returns %>% dplyr::filter(date >= TEST_START)

# Sector ETFs only (SPY is benchmark, not traded)
sector_tickers <- tickers[tickers != "SPY"]

sector_train <- ret_train %>% dplyr::filter(symbol %in% sector_tickers)
sector_test  <- ret_test  %>% dplyr::filter(symbol %in% sector_tickers)
spy_train    <- ret_train %>% dplyr::filter(symbol == "SPY")
spy_test     <- ret_test  %>% dplyr::filter(symbol == "SPY")
```

------------------------------------------------------------------------

## Executive Summary

This strategy uses K-Means clustering to classify eleven Select Sector SPDR ETFs into Cyclical and Defensive groups, then applies a Kalman Filter and SMA-based spread signal to rotate capital toward the cluster with positive momentum while stepping aside into cash when both clusters deteriorate simultaneously. Over the 24-year training period from June 2000 to December 2024, the optimized strategy produced superior risk-adjusted results relative to SPY, including a Sharpe Ratio of 0.906 versus 0.499, an Omega Ratio of 1.187 versus 1.106, and a maximum drawdown of 5.6% versus 55.2%, achieved by correctly reducing equity exposure during the 2001, 2008, and 2020 bear markets. In the 15-month out-of-sample testing period from January 2025 to March 2026, the strategy continued to lead all benchmarks on risk-adjusted metrics but trailed on absolute return, executing only one trade due to a signal threshold calibrated on 24 years of training data that proved too conservative for the low cross-sectional volatility environment of the test period.

------------------------------------------------------------------------

## 1. Intuitive Rationale

### Strategy Thesis

This strategy tests one hypothesis: that sector ETFs temporarily deviate from their co-moving peers and that those deviations are more likely to revert than persist. The mechanism being tested is whether unsupervised clustering on rolling return correlations can identify which ETFs genuinely move together, and whether overweighting the recent underperformer within each group generates risk-adjusted returns above a passive SPY benchmark. This project does not assume the hypothesis is correct. It tests it honestly, with full acknowledgment of where and why it fails.

### What Clustering Adds

A naive mean-reversion strategy would buy any ETF that has dropped recently. The problem is that not all ETFs are comparable: buying XLE because it underperformed XLK assumes those two sectors share a common driver, which they often do not. Clustering first identifies which ETFs are genuinely co-moving in the current market environment, based on 60 days of rolling return correlations. Only within those empirically identified peer groups does the mean-reversion signal have a structural basis. The cluster membership itself is dynamic: it updates monthly as correlations shift, so the strategy adapts to changing market structure rather than assuming fixed relationships.

### Why the Kalman Filter Is Required

Within-cluster underperformance can have two causes: temporary noise (the ETF will revert) or a genuine trend (the ETF is repricing to a new level). Buying into a downtrend on the assumption it will revert is the single largest failure mode of mean-reversion strategies. The Kalman Filter estimates the underlying trend in each ETF's price series by smoothing out daily noise. If the smoothed slope is negative beyond a threshold, the ETF is excluded from the portfolio entirely that day, regardless of how undervalued it appears relative to its cluster. The Kalman Filter is therefore not a signal enhancer; it is a risk gate that prevents the strategy from confusing structural repricing with temporary deviation.

### When This Strategy Fails

This strategy is designed for range-bound, mean-reverting environments and will fail in two specific conditions. The first is sustained directional trends: when sectors are repricing to fundamentally new levels (the 2022 rate shock driving XLU and XLRE down for months), within-cluster underperformers continue underperforming and the mean-reversion bet loses repeatedly. The Kalman gate reduces this exposure but does not eliminate it entirely, particularly at the onset of a trend before the slope signal has confirmed direction. The second is correlation collapse: during systemic shocks (March 2020, September 2008), all 11 ETFs correlate toward 1.0, the cluster structure becomes meaningless, and within-cluster z-scores carry no information. The strategy's cluster quality filter addresses this by disabling trading when average intra-cluster correlation falls below a minimum threshold.

### Academic Foundation

Two papers motivate the design. Ajayi (2025) demonstrates that unsupervised K-Means clustering identifies the same major market regimes as parametric Markov Switching models, validating that rolling clustering captures economically meaningful structural breaks without requiring distributional assumptions. Raju (2025) uses rolling K-Means on equity return data and finds that cluster-based signals generate statistically significant return differentials across regimes (Sharpe differential of 2 to 4 points, p \< 0.01), though those results are from Indian equities where return synchronicity is higher than in U.S. markets. Morck, Yeung, and Yu (2000) show that U.S. stocks have low return synchronicity relative to emerging markets, meaning firm-specific and sector-specific variation dominates macro signals. This implies that intra-cluster dispersion in U.S. sector ETFs will be meaningful and tradeable, but that Raju's Sharpe differentials represent an upper bound rather than a replication target.

------------------------------------------------------------------------

::: {.callout-note collapse="true"}
## Pre-Pivot Analysis

## 2. Data and Features

```{r data-sources-table}
#| message: false
#| warning: false
#| echo: false

tibble::tibble(
  Ticker      = c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
                  "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY"),
  Name        = c("SPDR S&P 500 ETF Trust", "Materials", "Communication Services",
                  "Energy", "Financials", "Industrials", "Technology",
                  "Consumer Staples", "Real Estate", "Health Care",
                  "Utilities", "Consumer Discretionary"),
  Role        = c("Benchmark", rep("Traded", 11)),
  `Available From` = c("2000-06-01", "2000-06-01", "2018-06-19", "2000-06-01",
                        "2000-06-01", "2000-06-01", "2000-06-01", "2000-06-01",
                        "2015-10-08", "2000-06-01", "2000-06-01", "2000-06-01"),
  Source      = rep("Yahoo Finance via tidyquant::tq_get()", 12)
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Universe: SPDR Sector ETFs and Benchmark",
    subtitle = "Daily adjusted close prices. Training: June 2000 to December 2024. Testing: January 2025 to March 2026."
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#f0f4ff"),
    locations = gt::cells_body(rows = Role == "Benchmark")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(columns = Ticker)
  ) %>%
  gt::tab_source_note("XLC available from June 2018; XLRE available from October 2015. Both included despite shorter histories.")
```

All price data is sourced via API with no local files. Log returns are computed as $r_t = \log(P_t / P_{t-1})$ on adjusted close prices, which account for dividends and splits. Log returns are used rather than simple returns because they are additive across time, approximately normally distributed, and produce a more stable correlation matrix, which is the direct input to the clustering algorithm. SPY is carried as the benchmark throughout but is never included in the traded universe or the clustering.

The clustering input is not the raw return series but the pairwise correlation matrix computed over a rolling 60-day window. Clustering directly on returns would group ETFs by recent absolute performance, crowding recent winners into one cluster regardless of whether they share underlying economic drivers. Correlation captures co-movement independent of magnitude: two ETFs are similar if they move in the same direction at the same time, not because they both went up. The 60-day window is a deliberate balance between responsiveness and stability. A shorter window (20 days) produces noisy correlations dominated by single events; a longer window (252 days) is too slow to reflect structural shifts in market relationships and would keep the portfolio anchored to stale cluster assignments. The 60-day parameter is treated as tunable and will be evaluated in the optimization section.

------------------------------------------------------------------------

## 3. Clustering

### 3.1 Cluster Hypothesis

The following cluster assignments are stated before any computation is performed. Results will be assessed honestly against these priors.

```{r cluster-hypothesis}
#| message: false
#| warning: false
#| echo: false

tibble::tibble(
  Cluster       = c("Cyclical", "Defensive", "Commodity", "Rate-Sensitive"),
  Sectors       = c("Industrials, Financials, Consumer Discretionary",
                    "Staples, Utilities, Healthcare",
                    "Energy, Materials",
                    "Technology, Communication Services, Real Estate"),
  `Core Driver` = c("GDP / growth",
                    "Stability / recession",
                    "Inflation / commodities",
                    "Interest rates")
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Cluster Hypothesis: Expected Groupings Before Computation",
    subtitle = "Stated prior to running the elbow method or hierarchical clustering."
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::cols_width(Cluster        ~ px(160),
                 Sectors        ~ px(300),
                 `Core Driver`  ~ px(200)) %>%
  gt::tab_source_note("Hypothesis informed by sector economic characteristics, not by training data.")
```

### 3.2 Selecting k: Elbow Method

```{r elbow-method}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 8
#| fig-height: 4

# Compute average correlation matrix over full training period
ret_wide_train <- sector_train %>%
  dplyr::select(date, symbol, ret) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = ret) %>%
  dplyr::arrange(date) %>%
  dplyr::select(-date)

cor_mat  <- cor(ret_wide_train, use = "pairwise.complete.obs")
dist_mat <- as.dist(1 - cor_mat)

# Hierarchical clustering
hc <- hclust(dist_mat, method = "ward.D2")

# WSS for k = 2 to 8
compute_wss <- function(k) {
  labels    <- cutree(hc, k = k)
  centroids <- tapply(seq_along(labels), labels, function(idx) idx)
  total_wss <- sum(sapply(unique(labels), function(cl) {
    members <- names(labels[labels == cl])
    if (length(members) < 2) return(0)
    sub_dist <- as.matrix(dist_mat)[members, members]
    sum(sub_dist^2) / (2 * length(members))
  }))
  tibble::tibble(k = k, wss = total_wss)
}

wss_df <- purrr::map_dfr(1:8, compute_wss)

ggplot(wss_df, aes(x = k, y = wss)) +
  geom_line(colour = "#2c7bb6", linewidth = 0.9) +
  geom_point(colour = "#2c7bb6", size = 3) +
  geom_vline(xintercept = 3, linetype = "dashed", colour = "#d7191c", linewidth = 0.8) +
  annotate("text", x = 3.2, y = max(wss_df$wss) * 0.95,
           label = "Selected k = 3", colour = "#d7191c", hjust = 0, size = 3.5) +
  scale_x_continuous(breaks = 1:8) +
  labs(
    title    = "Elbow Method: Within-Cluster Sum of Squares vs. Number of Clusters",
    subtitle = "Training period average correlation matrix. Ward.D2 hierarchical clustering.",
    x        = "Number of Clusters (k)",
    y        = "Within-Cluster Sum of Squares (WSS)",
    caption  = "Distance metric: d(i,j) = 1 - corr(i,j). Training period June 2000 to December 2024."
  ) +
  theme_minimal(base_size = 11)
```

Despite my initial hypothesis being k = 4, **the elbow I see is at k = 3.** There is a meaningful reduction in WSS from 3 -\> 4, but the decision to use k = 3 comes from the change in slope from 2 -\> 4 having a sharper angle than 3 -\> 5. There is no right answer here, but using 4 clusters creates the risk of weak clusters that change often, leading to increased transaction costs for the trading strategy.

### 3.3 Hierarchical Clustering

```{r hierarchical-clustering}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 9
#| fig-height: 5

K <- 3

# Map tickers to readable sector names
name_map <- c(
  XLB  = "Materials",       XLC  = "Communications",  XLE  = "Energy",
  XLF  = "Financials",      XLI  = "Industrials",     XLK  = "Technology",
  XLP  = "Consumer Staples", XLRE = "Real Estate",    XLV  = "Health Care",
  XLU  = "Utilities",       XLY  = "Consumer Disc."
)

hc_named         <- hc
hc_named$labels  <- name_map[hc$labels]

cluster_colors <- c("#2c7bb6", "#1a9641", "#d7191c")

factoextra::fviz_dend(
  hc_named,
  k                 = K,
  k_colors          = cluster_colors,
  color_labels_by_k = TRUE,
  rect              = TRUE,
  rect_fill         = TRUE,
  horiz             = TRUE,
  main              = "Hierarchical Clustering of Sector ETFs (Ward.D2, k = 3)",
  xlab              = "Height",
  ylab              = "",
  ggtheme           = theme_minimal(base_size = 11),
  labels_track_height = 0.8
) +
  labs(caption = "Distance: d(i,j) = 1 - corr(i,j). Training period June 2000 to December 2024.")
```

```{r cluster-membership}
#| message: false
#| warning: false
#| echo: false

cluster_assignments <- tibble::tibble(
  symbol  = names(cutree(hc, k = K)),
  cluster = cutree(hc, k = K)
) %>%
  dplyr::arrange(cluster, symbol)

# Silhouette score
sil       <- cluster::silhouette(cutree(hc, k = K), dist_mat)
avg_sil   <- round(mean(sil[, 3]), 3)

cluster_assignments %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    `ETFs`            = paste(symbol, collapse = ", "),
    `Size`            = dplyr::n(),
    .groups           = "drop"
  ) %>%
  dplyr::rename(Cluster = cluster) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Cluster Assignments at k = 3",
    subtitle = paste0("Average silhouette width: ", avg_sil, "  (> 0.25 indicates meaningful structure)")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_source_note("Assignments derived from full training period average correlation matrix.")
```

The dendrogram reveals which ETFs the algorithm groups together based on 24 years of rolling return correlations. ETFs that merge at low linkage distances are tightly co-moving; those that merge late are structurally more distinct. The red cut line at k = 3 produces three groups consistent with the prior hypothesis, with the commodity and rate-sensitive hypothesized clusters likely consolidating into one given their shared sensitivity to macro conditions. The average silhouette width quantifies cluster separation: values above 0.25 indicate the groupings are non-arbitrary, and values above 0.50 indicate strong structure. A low silhouette score here is a signal to revisit k before proceeding.

```{r sil-plot-k3}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 8
#| fig-height: 4

p_sil3 <- factoextra::fviz_silhouette(sil, print.summary = FALSE) +
  labs(title = paste0("k = 3  |  avg sil = ", avg_sil), x = NULL, y = "Silhouette Width") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```

### 3.4 Revisiting k: Testing k = 4

The silhouette width of 0.221 at k = 3 falls below the 0.25 threshold for meaningful cluster separation. The dendrogram suggests Energy is the likely culprit: it merges into the red cluster at the highest linkage distance, indicating it is structurally distinct from Industrials, Materials, and Financials but forced into the same group at k = 3. Splitting to k = 4 isolates Energy and tests whether cluster separation improves enough to justify the additional complexity and potential transaction costs from more frequent cluster reassignment.

```{r k4-test}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 9
#| fig-height: 5

K4 <- 4

hc_named4        <- hc
hc_named4$labels <- name_map[hc$labels]

factoextra::fviz_dend(
  hc_named4,
  k                 = K4,
  k_colors          = c("#2c7bb6", "#1a9641", "#d7191c", "#f4a736"),
  color_labels_by_k = TRUE,
  rect              = TRUE,
  rect_fill         = TRUE,
  horiz             = TRUE,
  main              = "Hierarchical Clustering of Sector ETFs (Ward.D2, k = 4)",
  xlab              = "Height",
  ylab              = "",
  ggtheme           = theme_minimal(base_size = 11),
  labels_track_height = 0.8
) +
  labs(caption = "Distance: d(i,j) = 1 - corr(i,j). Training period June 2000 to December 2024.")
```

```{r k4-silhouette}
#| message: false
#| warning: false
#| echo: false

sil4     <- cluster::silhouette(cutree(hc, k = K4), dist_mat)
avg_sil4 <- round(mean(sil4[, 3]), 3)

cluster_assignments4 <- tibble::tibble(
  symbol  = names(cutree(hc, k = K4)),
  cluster = cutree(hc, k = K4)
) %>%
  dplyr::arrange(cluster, symbol)

cluster_assignments4 %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    ETFs = paste(symbol, collapse = ", "),
    Size = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::rename(Cluster = cluster) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Cluster Assignments at k = 4",
    subtitle = paste0("Average silhouette width: ", avg_sil4, "  (> 0.25 indicates meaningful structure)")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_source_note("Assignments derived from full training period average correlation matrix.")
```

```{r sil-plot-k4}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 8
#| fig-height: 4

p_sil4 <- factoextra::fviz_silhouette(sil4, print.summary = FALSE) +
  labs(title = paste0("k = 4  |  avg sil = ", avg_sil4), x = NULL, y = "Silhouette Width") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```


### 3.5 Revisiting k: Testing k = 5

At k = 4, Real Estate and Utilities split off before Energy, which remains grouped with Industrials, Materials, and Financials. The silhouette plot reveals that these two small clusters (size 2 each) are the least well-separated. Testing k = 5 allows Energy to isolate into its own cluster, which is the structural behavior suggested by the dendrogram throughout -- Energy consistently merges last at the highest linkage distance, indicating its commodity-driven return profile is genuinely distinct from all other sectors.

```{r k5-test}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 9
#| fig-height: 5

K5 <- 5

hc_named5        <- hc
hc_named5$labels <- name_map[hc$labels]

factoextra::fviz_dend(
  hc_named5,
  k                   = K5,
  k_colors            = c("#2c7bb6", "#1a9641", "#d7191c", "#f4a736", "#984ea3"),
  color_labels_by_k   = TRUE,
  rect                = TRUE,
  rect_fill           = TRUE,
  horiz               = TRUE,
  main                = "Hierarchical Clustering of Sector ETFs (Ward.D2, k = 5)",
  xlab                = "Height",
  ylab                = "",
  ggtheme             = theme_minimal(base_size = 11),
  labels_track_height = 0.8
) +
  labs(caption = "Distance: d(i,j) = 1 - corr(i,j). Training period June 2000 to December 2024.")
```

```{r k5-silhouette}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 8
#| fig-height: 4

K5         <- 5
sil5       <- cluster::silhouette(cutree(hc, k = K5), dist_mat)
avg_sil5   <- round(mean(sil5[, 3]), 3)

cluster_assignments5 <- tibble::tibble(
  symbol  = names(cutree(hc, k = K5)),
  cluster = cutree(hc, k = K5)
) %>%
  dplyr::arrange(cluster, symbol)

cluster_assignments5 %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    ETFs = paste(symbol, collapse = ", "),
    Size = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::rename(Cluster = cluster) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Cluster Assignments at k = 5",
    subtitle = paste0("Average silhouette width: ", avg_sil5, "  (> 0.50 indicates reasonable structure)")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_source_note("Assignments derived from full training period average correlation matrix.")

p_sil5 <- factoextra::fviz_silhouette(sil5, print.summary = FALSE) +
  labs(title = paste0("k = 5  |  avg sil = ", avg_sil5), x = NULL, y = "Silhouette Width") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```

### 3.6 Revisiting k: Testing k = 6

```{r k6-test}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 9
#| fig-height: 5

K6 <- 6

hc_named6        <- hc
hc_named6$labels <- name_map[hc$labels]

factoextra::fviz_dend(
  hc_named6,
  k                   = K6,
  k_colors            = c("#2c7bb6", "#1a9641", "#d7191c", "#f4a736", "#984ea3", "#e41a1c"),
  color_labels_by_k   = TRUE,
  rect                = TRUE,
  rect_fill           = TRUE,
  horiz               = TRUE,
  main                = "Hierarchical Clustering of Sector ETFs (Ward.D2, k = 6)",
  xlab                = "Height",
  ylab                = "",
  ggtheme             = theme_minimal(base_size = 11),
  labels_track_height = 0.8
) +
  labs(caption = "Distance: d(i,j) = 1 - corr(i,j). Training period June 2000 to December 2024.")
```

```{r k6-silhouette}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 8
#| fig-height: 4

sil6     <- cluster::silhouette(cutree(hc, k = K6), dist_mat)
avg_sil6 <- round(mean(sil6[, 3]), 3)

cluster_assignments6 <- tibble::tibble(
  symbol  = names(cutree(hc, k = K6)),
  cluster = cutree(hc, k = K6)
) %>%
  dplyr::arrange(cluster, symbol)

cluster_assignments6 %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    ETFs = paste(symbol, collapse = ", "),
    Size = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::rename(Cluster = cluster) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Cluster Assignments at k = 6",
    subtitle = paste0("Average silhouette width: ", avg_sil6, "  (> 0.50 indicates reasonable structure)")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_source_note("Assignments derived from full training period average correlation matrix.")

p_sil6 <- factoextra::fviz_silhouette(sil6, print.summary = FALSE) +
  labs(title = paste0("k = 6  |  avg sil = ", avg_sil6), x = NULL, y = "Silhouette Width") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")
```

### 3.7 Silhouette Comparison: k = 3 to 6

```{r sil-comparison}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 7

(p_sil3 | p_sil4) / (p_sil5 | p_sil6) +
  patchwork::plot_annotation(
    title   = "Silhouette Width Comparison Across k",
    subtitle = "Higher and more uniform bars = better cluster separation. Dashed line = cluster average.",
    caption  = "Training period June 2000 to December 2024. Ward.D2 hierarchical clustering."
  )
```


### 3.8 Cluster Stability Over Time

```{r rolling-cluster-stability}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 11
#| fig-height: 5

K_USE <- 3
LOOKBACK <- 60

# Month-end dates with at least LOOKBACK trading days of history
all_dates <- sector_train %>%
  dplyr::filter(symbol == "XLK") %>%
  dplyr::arrange(date) %>%
  dplyr::pull(date)

month_ends <- sector_train %>%
  dplyr::filter(symbol == "XLK") %>%
  dplyr::mutate(ym = lubridate::floor_date(date, "month")) %>%
  dplyr::group_by(ym) %>%
  dplyr::slice_max(date, n = 1) %>%
  dplyr::ungroup() %>%
  dplyr::filter(date >= all_dates[LOOKBACK]) %>%
  dplyr::pull(date)

# Compute raw cluster assignments for one window
cluster_one_window <- function(end_date) {
  window <- sector_train %>%
    dplyr::filter(date <= end_date) %>%
    dplyr::group_by(symbol) %>%
    dplyr::slice_tail(n = LOOKBACK) %>%
    dplyr::ungroup() %>%
    dplyr::select(date, symbol, ret) %>%
    tidyr::pivot_wider(names_from = symbol, values_from = ret) %>%
    dplyr::select(-date)

  cm   <- cor(window, use = "pairwise.complete.obs")
  dm   <- as.dist(1 - cm)
  hc_w <- hclust(dm, method = "ward.D2")
  labs <- cutree(hc_w, k = K_USE)
  tibble::tibble(date = end_date, symbol = names(labs), raw_cluster = as.integer(labs))
}

raw_clusters <- purrr::map_dfr(month_ends, cluster_one_window)

# Relabel clusters across windows using maximum-overlap greedy matching
relabel_rolling <- function(raw, k) {
  windows <- sort(unique(raw$date))

  first <- raw %>%
    dplyr::filter(date == windows[1]) %>%
    dplyr::mutate(cluster = raw_cluster)

  out      <- first
  prev_map <- first %>% dplyr::select(symbol, cluster) %>% tibble::deframe()

  for (i in seq(2, length(windows))) {
    curr <- raw %>% dplyr::filter(date == windows[i])
    curr_map <- curr %>% dplyr::select(symbol, raw_cluster) %>% tibble::deframe()

    # Build k x k overlap matrix: rows = old clusters, cols = new clusters
    overlap <- outer(
      1:k, 1:k,
      Vectorize(function(old_c, new_c) {
        length(intersect(names(prev_map[prev_map == old_c]),
                         names(curr_map[curr_map == new_c])))
      })
    )

    # Greedy one-to-one assignment
    mapping   <- integer(k)
    avail_old <- 1:k
    avail_new <- 1:k

    for (iter in 1:k) {
      sub      <- overlap[avail_old, avail_new, drop = FALSE]
      max_pos  <- which(sub == max(sub), arr.ind = TRUE)[1, ]
      old_c    <- avail_old[max_pos[1]]
      new_c    <- avail_new[max_pos[2]]
      mapping[new_c] <- old_c
      avail_old <- avail_old[avail_old != old_c]
      avail_new <- avail_new[avail_new != new_c]
    }

    relabeled <- curr %>%
      dplyr::mutate(cluster = mapping[raw_cluster])

    out      <- dplyr::bind_rows(out, relabeled)
    prev_map <- relabeled %>% dplyr::select(symbol, cluster) %>% tibble::deframe()
  }
  out
}

stable_clusters <- relabel_rolling(raw_clusters, K_USE)

# Sector name map for display
name_map_vec <- c(
  XLB = "Materials", XLC = "Communications", XLE = "Energy",
  XLF = "Financials", XLI = "Industrials",   XLK = "Technology",
  XLP = "Cons. Staples", XLRE = "Real Estate", XLV = "Health Care",
  XLU = "Utilities", XLY = "Cons. Disc."
)

heatmap_df <- stable_clusters %>%
  dplyr::mutate(
    sector  = name_map_vec[symbol],
    cluster = factor(cluster)
  )

ggplot(heatmap_df, aes(x = date, y = sector, fill = cluster)) +
  geom_tile(colour = "white", linewidth = 0.15) +
  scale_fill_manual(
    values = c("1" = "#2c7bb6", "2" = "#1a9641", "3" = "#d7191c"),
    name   = "Cluster"
  ) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title   = "Rolling Cluster Assignments Over Time (k = 3, 60-Day Window)",
    subtitle = "Each cell = cluster assigned at that month-end. Labels matched across windows by maximum membership overlap.",
    x       = NULL,
    y       = NULL,
    caption = "Training period June 2000 to December 2024. Monthly rebalancing frequency."
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(),
        axis.text.y = element_text(size = 9))
```

```{r cluster-stability-metrics}
#| message: false
#| warning: false
#| echo: false

# Window-to-window stability: % ETFs unchanged
stability_ww <- stable_clusters %>%
  dplyr::arrange(symbol, date) %>%
  dplyr::group_by(symbol) %>%
  dplyr::mutate(changed = cluster != dplyr::lag(cluster)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(!is.na(changed)) %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(pct_stable = mean(!changed), .groups = "drop")

avg_stability <- round(mean(stability_ww$pct_stable) * 100, 1)

# Per-ETF persistence: % of windows each ETF stays in its modal cluster
per_etf <- stable_clusters %>%
  dplyr::group_by(symbol) %>%
  dplyr::summarise(
    modal_cluster   = as.integer(names(which.max(table(cluster)))),
    pct_in_modal    = round(mean(cluster == modal_cluster) * 100, 1),
    n_switches      = sum(cluster != dplyr::lag(cluster), na.rm = TRUE),
    .groups         = "drop"
  ) %>%
  dplyr::mutate(sector = name_map_vec[symbol]) %>%
  dplyr::select(sector, modal_cluster, pct_in_modal, n_switches) %>%
  dplyr::arrange(dplyr::desc(pct_in_modal))

per_etf %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Per-ETF Cluster Persistence",
    subtitle = paste0("Average window-to-window stability: ", avg_stability,
                      "% of ETFs unchanged each month")
  ) %>%
  gt::cols_label(
    sector        = "Sector",
    modal_cluster = "Primary Cluster",
    pct_in_modal  = "Time in Primary Cluster (%)",
    n_switches    = "Cluster Switches"
  ) %>%
  gt::fmt_number(columns = pct_in_modal, decimals = 1) %>%
  gt::data_color(
    columns = pct_in_modal,
    palette = c("#d7191c", "white", "#1a9641"),
    domain  = c(50, 100)
  ) %>%
  gt::tab_source_note("Modal cluster = cluster assigned most frequently over training period.")
```

The heatmap reveals that k = 3 with rolling return correlations alone produces highly unstable cluster assignments. Most ETFs switch clusters almost every month with no persistent color pattern, which means the strategy would be rebalancing into a different cluster structure nearly every period. This is a problem: mean reversion within a cluster requires that cluster to be stable long enough for the z-score signal to converge and the position to recover. Constant reshuffling generates transaction costs without giving the signal time to play out. The per-ETF stability table quantifies this -- sectors that theory predicts should be permanently defensive (Utilities, Consumer Staples, Health Care) are not consistently grouped together. Rolling return correlations alone are too regime-sensitive to serve as a reliable clustering input.

### 3.9 Pivot: Adding Rolling Volatility as a Second Feature

The instability finding above is not a reason to abandon clustering. It is a reason to enrich the feature set. Rolling return correlations capture the direction of co-movement but not the character of each ETF's risk profile. A defensive ETF and a cyclical ETF can temporarily correlate during a broad market rally, causing the algorithm to group them together even though their structural characters are completely different. Rolling realized volatility is a structural feature that is far more persistent: Utilities and Consumer Staples are low-vol in every market environment; Energy and Technology are high-vol across regimes. Adding volatility as a second clustering input gives the algorithm a stable anchor that persists through correlation regime shifts.

The implementation combines two normalized distance matrices: the correlation distance (1 minus pairwise correlation, same as above) and a volatility distance (absolute difference in 60-day realized volatility between each ETF pair, normalized to a 0 to 1 scale). The combined distance is a 50/50 weighted average of the two, applied to the same Ward.D2 hierarchical clustering. The weight is a parameter that will be evaluated in the optimization section.

```{r vol-feature-clustering}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 11
#| fig-height: 5

VOL_WEIGHT <- 0.5  # weight on volatility distance (1 - VOL_WEIGHT on correlation)

cluster_one_window_v2 <- function(end_date) {
  window <- sector_train %>%
    dplyr::filter(date <= end_date) %>%
    dplyr::group_by(symbol) %>%
    dplyr::slice_tail(n = LOOKBACK) %>%
    dplyr::ungroup()

  ret_wide <- window %>%
    dplyr::select(date, symbol, ret) %>%
    tidyr::pivot_wider(names_from = symbol, values_from = ret) %>%
    dplyr::select(-date)

  # Correlation distance
  cm       <- cor(ret_wide, use = "pairwise.complete.obs")
  dist_cor <- as.matrix(1 - cm)

  # Volatility distance: normalized absolute difference in realized vol
  etf_vols  <- apply(ret_wide, 2, sd, na.rm = TRUE) * sqrt(252)
  vol_diff  <- as.matrix(dist(etf_vols, method = "euclidean"))
  vol_range <- max(vol_diff)
  dist_vol  <- if (vol_range > 0) vol_diff / vol_range else vol_diff

  # Combined distance
  dist_combined <- as.dist((1 - VOL_WEIGHT) * dist_cor + VOL_WEIGHT * dist_vol)

  hc_v2 <- hclust(dist_combined, method = "ward.D2")
  labs  <- cutree(hc_v2, k = K_USE)

  tibble::tibble(date = end_date, symbol = names(labs), raw_cluster = as.integer(labs))
}

raw_v2     <- purrr::map_dfr(month_ends, cluster_one_window_v2)
stable_v2  <- relabel_rolling(raw_v2, K_USE)

heatmap_v2 <- stable_v2 %>%
  dplyr::mutate(sector  = name_map_vec[symbol],
                cluster = factor(cluster))

ggplot(heatmap_v2, aes(x = date, y = sector, fill = cluster)) +
  geom_tile(colour = "white", linewidth = 0.15) +
  scale_fill_manual(
    values = c("1" = "#2c7bb6", "2" = "#1a9641", "3" = "#d7191c"),
    name   = "Cluster"
  ) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "Rolling Cluster Assignments: Correlation + Volatility Features (k = 3)",
    subtitle = "Combined distance = 50% correlation distance + 50% volatility distance.",
    x        = NULL,
    y        = NULL,
    caption  = "Training period June 2000 to December 2024. Monthly rebalancing frequency."
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid    = element_blank(),
        axis.text.y   = element_text(size = 9))
```

```{r vol-stability-metrics}
#| message: false
#| warning: false
#| echo: false

stability_v2 <- stable_v2 %>%
  dplyr::arrange(symbol, date) %>%
  dplyr::group_by(symbol) %>%
  dplyr::mutate(changed = cluster != dplyr::lag(cluster)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(!is.na(changed)) %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(pct_stable = mean(!changed), .groups = "drop")

avg_stability_v2 <- round(mean(stability_v2$pct_stable) * 100, 1)

per_etf_v2 <- stable_v2 %>%
  dplyr::group_by(symbol) %>%
  dplyr::summarise(
    modal_cluster = as.integer(names(which.max(table(cluster)))),
    pct_in_modal  = round(mean(cluster == modal_cluster) * 100, 1),
    n_switches    = sum(cluster != dplyr::lag(cluster), na.rm = TRUE),
    .groups       = "drop"
  ) %>%
  dplyr::mutate(sector = name_map_vec[symbol]) %>%
  dplyr::select(sector, modal_cluster, pct_in_modal, n_switches) %>%
  dplyr::arrange(dplyr::desc(pct_in_modal))

per_etf_v2 %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Per-ETF Cluster Persistence: Correlation + Volatility Features",
    subtitle = paste0("Average window-to-window stability: ", avg_stability_v2,
                      "% of ETFs unchanged each month")
  ) %>%
  gt::cols_label(
    sector        = "Sector",
    modal_cluster = "Primary Cluster",
    pct_in_modal  = "Time in Primary Cluster (%)",
    n_switches    = "Cluster Switches"
  ) %>%
  gt::fmt_number(columns = pct_in_modal, decimals = 1) %>%
  gt::data_color(
    columns = pct_in_modal,
    palette = c("#d7191c", "white", "#1a9641"),
    domain  = c(50, 100)
  ) %>%
  gt::tab_source_note("Compare to correlation-only table above. Improvement in pct_in_modal and fewer switches indicates feature enrichment is working.")
```

*\[Compare heatmap and stability metrics to correlation-only version above. If persistence improves materially, adopt correlation + volatility as the clustering input going forward.\]*

:::

------------------------------------------------------------------------

## 2. Data and Features

```{r universe-table}
#| message: false
#| warning: false
#| echo: false

tibble::tibble(
  Ticker = c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY"),
  Name   = c("SPDR S&P 500 ETF Trust", "Materials", "Communication Services",
             "Energy", "Financials", "Industrials", "Technology",
             "Consumer Staples", "Real Estate", "Health Care",
             "Utilities", "Consumer Discretionary"),
  Role   = c("Benchmark", rep("Traded", 11)),
  `Available From` = c(
    "2000-06-01", "2000-06-01", "2018-06-19", "2000-06-01",
    "2000-06-01", "2000-06-01", "2000-06-01", "2000-06-01",
    "2015-10-08", "2000-06-01", "2000-06-01", "2000-06-01"),
  Source = rep("Yahoo Finance via tidyquant::tq_get()", 12)
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Universe: SPDR Sector ETFs and Benchmark",
    subtitle = "Training: June 2000 to December 2024. Testing: January 2025 to March 2026."
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#f0f4ff"),
    locations = gt::cells_body(rows = Role == "Benchmark")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(columns = Ticker)
  ) %>%
  gt::tab_source_note("XLC available from June 2018; XLRE from October 2015. Both included despite shorter histories.")
```

All price data is sourced via API with no local files. Log returns are computed as $r_t = \log(P_t / P_{t-1})$ on adjusted close prices, which account for dividends and splits. The training period runs through December 2024 and is the only window used to fit any model parameter; the testing period from January 2025 onward is strictly held out. SPY is carried as the benchmark throughout but is never included in the traded universe or the clustering.

The clustering input in this revised approach is a feature matrix built from each ETF's relationship to SPY, not pairwise correlations between sector ETFs. This addresses the core instability problem identified in the original approach: correlations between sector ETFs are highly regime-sensitive and shift sharply during stress events, causing cluster assignments to flip frequently. SPY-relative features measure each ETF's structural character (how much it outperforms the market, how tightly it tracks it, and how it behaves during drawdowns), which are more persistent properties across market cycles. All three features share a single rolling window so the strategy has one tunable parameter rather than three independent ones.

```{r features-table}
#| message: false
#| warning: false
#| echo: false

tibble::tibble(
  Feature   = c("Cumulative Excess Return", "Beta", "Max Drawdown Ratio"),
  Formula   = c(
    "sum(r_ETF) - sum(r_SPY) over rolling window",
    "cov(r_ETF, r_SPY) / var(r_SPY) over rolling window",
    "MaxDD(ETF) / MaxDD(SPY) over rolling window"
  ),
  `What It Captures` = c(
    "Has this ETF been outperforming or underperforming the market recently?",
    "How sensitive is this ETF to broad market moves? High beta vs. defensive character.",
    "Does this ETF suffer deeper drawdowns than the market? Tail risk character."
  ),
  `Rolling Window` = rep("Same tunable parameter (default: 60 days)", 3)
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Clustering Features: SPY-Relative Characteristics",
    subtitle = "All features computed over the same rolling window. Normalized (z-score scaled) before clustering."
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::cols_width(
    Feature              ~ px(180),
    Formula              ~ px(260),
    `What It Captures`   ~ px(280),
    `Rolling Window`     ~ px(200)
  ) %>%
  gt::tab_source_note("Features are z-score normalized before Euclidean distance is computed. Single rolling window = single optimization parameter.")
```

Normalizing all three features before clustering is essential because they operate on different scales: excess return is a small decimal, beta is centred near 1.0, and the drawdown ratio has no natural upper bound. Z-score scaling ensures no single feature dominates the distance matrix by virtue of its units. Max Drawdown Ratio is defined as the ETF's peak-to-trough decline divided by SPY's peak-to-trough decline over the same window, so a ratio above 1.0 means the ETF experienced a deeper drawdown than the market, a signature of cyclical, high-beta sectors. A ratio below 1.0 indicates defensive character. This gives the algorithm a tail-risk anchor that is structurally persistent even when short-run correlations are noisy.

------------------------------------------------------------------------

## 3. Clustering

### 3.1 Selecting k: Elbow Method

```{r spy-features-elbow}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 8
#| fig-height: 4

LOOKBACK_ELBOW <- 60   # default window; same parameter used throughout

# Helper: max drawdown from a return vector
max_dd <- function(r) {
  prices <- cumprod(1 + r)
  peak   <- cummax(prices)
  min((prices - peak) / peak)
}

# Use the most recent LOOKBACK_ELBOW days of training data for static k selection
last_date <- max(ret_train$date)

window_train <- ret_train %>%
  dplyr::filter(symbol %in% sector_tickers) %>%
  dplyr::group_by(symbol) %>%
  dplyr::slice_tail(n = LOOKBACK_ELBOW) %>%
  dplyr::ungroup()

spy_window <- ret_train %>%
  dplyr::filter(symbol == "SPY") %>%
  dplyr::slice_tail(n = LOOKBACK_ELBOW)

spy_cum_ret  <- sum(spy_window$ret, na.rm = TRUE)
spy_var      <- var(spy_window$ret, na.rm = TRUE)
spy_mdd      <- max_dd(spy_window$ret)

feature_mat <- window_train %>%
  dplyr::group_by(symbol) %>%
  dplyr::summarise(
    excess_ret = sum(ret, na.rm = TRUE) - spy_cum_ret,
    beta       = cov(ret, spy_window$ret[seq_len(dplyr::n())], use = "complete.obs") / spy_var,
    mdd_ratio  = max_dd(ret) / spy_mdd,
    .groups    = "drop"
  )

# Normalize (z-score scale) before clustering
feat_scaled <- feature_mat %>%
  dplyr::select(excess_ret, beta, mdd_ratio) %>%
  scale()

rownames(feat_scaled) <- feature_mat$symbol

dist_spy <- dist(feat_scaled, method = "euclidean")
hc_spy   <- hclust(dist_spy, method = "ward.D2")

# WSS for k = 1:8
compute_wss_spy <- function(k) {
  if (k == 1) {
    total_wss <- sum(as.matrix(dist_spy)^2) / (2 * nrow(feat_scaled))
    return(tibble::tibble(k = k, wss = total_wss))
  }
  labs <- cutree(hc_spy, k = k)
  total_wss <- sum(sapply(unique(labs), function(cl) {
    members <- which(labs == cl)
    if (length(members) < 2) return(0)
    sub_d <- as.matrix(dist_spy)[members, members]
    sum(sub_d^2) / (2 * length(members))
  }))
  tibble::tibble(k = k, wss = total_wss)
}

wss_spy <- purrr::map_dfr(1:8, compute_wss_spy)

ggplot(wss_spy, aes(x = k, y = wss)) +
  geom_line(colour = "#2c7bb6", linewidth = 0.9) +
  geom_point(colour = "#2c7bb6", size = 3) +
  geom_vline(xintercept = 2, linetype = "dashed", colour = "#d7191c", linewidth = 0.8) +
  annotate("text", x = 2.15, y = max(wss_spy$wss) * 0.95,
           label = "Selected k = 2", colour = "#d7191c", hjust = 0, size = 3.5) +
  scale_x_continuous(breaks = 1:8) +
  labs(
    title    = "Elbow Method: Within-Cluster Sum of Squares vs. Number of Clusters",
    subtitle = "SPY-relative features: cumulative excess return, beta, max drawdown ratio. Ward.D2, Euclidean distance.",
    x        = "Number of Clusters (k)",
    y        = "Within-Cluster Sum of Squares (WSS)",
    caption  = paste0("Features z-score normalized. Last ", LOOKBACK_ELBOW, " days of training data. June 2000 to December 2024.")
  ) +
  theme_minimal(base_size = 11)
```

The elbow is unambiguous at k = 2. The drop from k = 1 to k = 2 is roughly four times larger than the drop from k = 2 to k = 3, and the curve is nearly flat beyond that. With three SPY-relative features measuring a fundamentally one-dimensional spectrum from defensive to cyclical, two clusters captures the dominant structure without forcing a third group that the features do not clearly support.

**Note for signal generation:** k = 2 produces one cluster that tracks the market closely and one that deviates from it. The deviating cluster can sit on either side of SPY: it may be the group currently outperforming the market or the group currently underperforming. The cumulative excess return feature carries the sign of that deviation, so the signal generation layer must distinguish direction within the cluster rather than assuming the "different" cluster is always the buy or always the sell.

### 3.2 Hierarchical Clustering: k = 2 vs k = 3

```{r spy-dendrograms}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 9

name_map_spy <- c(
  XLB = "Materials",      XLC = "Communications", XLE = "Energy",
  XLF = "Financials",     XLI = "Industrials",    XLK = "Technology",
  XLP = "Cons. Staples",  XLRE = "Real Estate",   XLV = "Health Care",
  XLU = "Utilities",      XLY = "Cons. Disc."
)

hc_spy_named         <- hc_spy
hc_spy_named$labels  <- name_map_spy[hc_spy$labels]

p_dend2 <- factoextra::fviz_dend(
  hc_spy_named,
  k                   = 2,
  k_colors            = c("#2c7bb6", "#d7191c"),
  color_labels_by_k   = TRUE,
  rect                = TRUE,
  rect_fill           = TRUE,
  horiz               = TRUE,
  main                = "k = 2",
  xlab                = "Height",
  ylab                = "",
  ggtheme             = theme_minimal(base_size = 10),
  labels_track_height = 0.8
) +
  labs(caption = "SPY-relative features, Ward.D2, Euclidean distance.")

p_dend3 <- factoextra::fviz_dend(
  hc_spy_named,
  k                   = 3,
  k_colors            = c("#2c7bb6", "#1a9641", "#d7191c"),
  color_labels_by_k   = TRUE,
  rect                = TRUE,
  rect_fill           = TRUE,
  horiz               = TRUE,
  main                = "k = 3",
  xlab                = "Height",
  ylab                = "",
  ggtheme             = theme_minimal(base_size = 10),
  labels_track_height = 0.8
)

p_dend2 / p_dend3 +
  patchwork::plot_annotation(
    title    = "Hierarchical Clustering of Sector ETFs (Ward.D2)",
    subtitle = "Features: cumulative excess return vs SPY, beta, max drawdown ratio. All z-score normalized.",
    caption  = "Training period June 2000 to December 2024."
  )
```

```{r spy-silhouettes}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 5

sil2     <- cluster::silhouette(cutree(hc_spy, k = 2), dist_spy)
sil3_spy <- cluster::silhouette(cutree(hc_spy, k = 3), dist_spy)

avg_sil2     <- round(mean(sil2[, 3]), 3)
avg_sil3_spy <- round(mean(sil3_spy[, 3]), 3)

p_s2 <- factoextra::fviz_silhouette(sil2, print.summary = FALSE) +
  labs(title = paste0("k = 2  |  avg sil = ", avg_sil2), x = NULL, y = "Silhouette Width") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")

p_s3 <- factoextra::fviz_silhouette(sil3_spy, print.summary = FALSE) +
  labs(title = paste0("k = 3  |  avg sil = ", avg_sil3_spy), x = NULL, y = "Silhouette Width") +
  theme_minimal(base_size = 9) +
  theme(legend.position = "none")

p_s2 | p_s3
```

```{r spy-cluster-table}
#| message: false
#| warning: false
#| echo: false

K_FINAL <- 2

cluster_assignments_spy <- tibble::tibble(
  symbol  = rownames(feat_scaled),
  cluster = cutree(hc_spy, k = K_FINAL)
) %>%
  dplyr::arrange(cluster, symbol) %>%
  dplyr::mutate(sector = name_map_spy[symbol])

cluster_assignments_spy %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(
    Sectors = paste(sector, collapse = ", "),
    Size    = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::rename(Cluster = cluster) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Cluster Assignments at k = 2",
    subtitle = paste0("Average silhouette width: ", avg_sil2,
                      "  (benchmark: > 0.50 is reasonable per class notes)")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_source_note("Assignments derived from last 60 days of training data. Labels will be revisited in rolling stability analysis.")
```

k = 2 is confirmed. Silhouette width of 0.521 clears the 0.50 benchmark from class notes and is materially above the 0.457 achieved at k = 3. Cluster 1 groups the defensive and commodity-sensitive sectors (Materials, Energy, Consumer Staples, Real Estate, Utilities, Health Care) and Cluster 2 groups the higher-beta, growth-oriented sectors (Communications, Financials, Industrials, Technology, Consumer Discretionary). This split is economically coherent: it maps directly onto the classic defensive versus cyclical distinction that drives sector rotation in practice.

### 3.3 Cluster Stability Over Time

Rather than relying on greedy overlap matching to track labels across rolling windows, this analysis anchors labels by economic characteristic. Each month, whichever cluster has the higher mean beta is assigned the label Cyclical; the other is assigned Defensive. Beta is a structural property that is far more persistent than pairwise correlations, so this approach produces consistent labels without depending on the arbitrary order that `hclust` assigns cluster numbers.

```{r spy-rolling-stability}
#| message: false
#| warning: false
#| echo: false
#| cache: true
#| fig-width: 11
#| fig-height: 5

LOOKBACK_ROLL <- 60

all_dates_roll <- sector_train %>%
  dplyr::filter(symbol == "XLK") %>%
  dplyr::arrange(date) %>%
  dplyr::pull(date)

month_ends_spy <- sector_train %>%
  dplyr::filter(symbol == "XLK") %>%
  dplyr::mutate(ym = lubridate::floor_date(date, "month")) %>%
  dplyr::group_by(ym) %>%
  dplyr::slice_max(date, n = 1) %>%
  dplyr::ungroup() %>%
  dplyr::filter(date >= all_dates_roll[LOOKBACK_ROLL]) %>%
  dplyr::pull(date)

cluster_window_spy <- function(end_date) {
  win_sec <- sector_train %>%
    dplyr::filter(date <= end_date) %>%
    dplyr::group_by(symbol) %>%
    dplyr::slice_tail(n = LOOKBACK_ROLL) %>%
    dplyr::ungroup()

  win_spy <- spy_train %>%
    dplyr::filter(date <= end_date) %>%
    dplyr::slice_tail(n = LOOKBACK_ROLL)

  spy_cr  <- sum(win_spy$ret, na.rm = TRUE)
  spy_v   <- var(win_spy$ret, na.rm = TRUE)
  spy_md  <- max_dd(win_spy$ret)

  feats <- win_sec %>%
    dplyr::group_by(symbol) %>%
    dplyr::summarise(
      excess_ret = sum(ret, na.rm = TRUE) - spy_cr,
      beta       = cov(ret, win_spy$ret[seq_len(dplyr::n())],
                       use = "complete.obs") / spy_v,
      mdd_ratio  = max_dd(ret) / spy_md,
      .groups    = "drop"
    )

  mat <- feats %>%
    dplyr::select(excess_ret, beta, mdd_ratio) %>%
    scale()

  rownames(mat) <- feats$symbol

  hc_w  <- hclust(dist(mat, method = "euclidean"), method = "ward.D2")
  labs  <- cutree(hc_w, k = 2)

  # Anchor label by beta: higher mean beta = Cyclical
  mean_beta <- tapply(feats$beta, labs, mean)
  cyclical_raw <- as.integer(names(which.max(mean_beta)))

  tibble::tibble(
    date    = end_date,
    symbol  = feats$symbol,
    cluster = ifelse(labs == cyclical_raw, "Cyclical", "Defensive")
  )
}

rolling_spy <- purrr::map_dfr(month_ends_spy, cluster_window_spy)

name_map_roll <- c(
  XLB = "Materials",     XLC = "Communications", XLE = "Energy",
  XLF = "Financials",    XLI = "Industrials",    XLK = "Technology",
  XLP = "Cons. Staples", XLRE = "Real Estate",   XLV = "Health Care",
  XLU = "Utilities",     XLY = "Cons. Disc."
)

heatmap_spy <- rolling_spy %>%
  dplyr::mutate(sector = name_map_roll[symbol])

ggplot(heatmap_spy, aes(x = date, y = sector, fill = cluster)) +
  geom_tile(colour = "white", linewidth = 0.15) +
  scale_fill_manual(
    values = c("Cyclical" = "#d7191c", "Defensive" = "#2c7bb6"),
    name   = "Cluster"
  ) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(
    title    = "Rolling Cluster Assignments Over Time (k = 2, 60-Day Window)",
    subtitle = "Labels anchored by beta: higher mean beta cluster = Cyclical each month.",
    x        = NULL,
    y        = NULL,
    caption  = "Training period June 2000 to December 2024. Monthly rebalancing frequency."
  ) +
  theme_minimal(base_size = 11) +
  theme(panel.grid   = element_blank(),
        axis.text.y  = element_text(size = 9))
```

```{r spy-stability-table}
#| message: false
#| warning: false
#| echo: false

stability_spy <- rolling_spy %>%
  dplyr::arrange(symbol, date) %>%
  dplyr::group_by(symbol) %>%
  dplyr::mutate(changed = cluster != dplyr::lag(cluster)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(!is.na(changed)) %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(pct_stable = mean(!changed), .groups = "drop")

avg_stability_spy <- round(mean(stability_spy$pct_stable) * 100, 1)

per_etf_spy <- rolling_spy %>%
  dplyr::group_by(symbol) %>%
  dplyr::summarise(
    modal_cluster = names(which.max(table(cluster))),
    pct_in_modal  = round(mean(cluster == modal_cluster) * 100, 1),
    n_switches    = sum(cluster != dplyr::lag(cluster), na.rm = TRUE),
    .groups       = "drop"
  ) %>%
  dplyr::mutate(sector = name_map_roll[symbol]) %>%
  dplyr::select(sector, modal_cluster, pct_in_modal, n_switches) %>%
  dplyr::arrange(dplyr::desc(pct_in_modal))

per_etf_spy %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Per-ETF Cluster Persistence (k = 2, Beta-Anchored Labels)",
    subtitle = paste0("Average window-to-window stability: ", avg_stability_spy,
                      "% of ETFs unchanged each month")
  ) %>%
  gt::cols_label(
    sector        = "Sector",
    modal_cluster = "Primary Cluster",
    pct_in_modal  = "Time in Primary Cluster (%)",
    n_switches    = "Cluster Switches"
  ) %>%
  gt::fmt_number(columns = pct_in_modal, decimals = 1) %>%
  gt::data_color(
    columns = pct_in_modal,
    palette = c("#d7191c", "white", "#2c7bb6"),
    domain  = c(50, 100)
  ) %>%
  gt::tab_source_note("Modal cluster = label assigned most frequently over training period. Defensive = blue, Cyclical = red.")
```

This approach is a material improvement over the correlation-only method used before the pivot. Three differences stand out. First, **silhouette width increased from roughly 0.22 to 0.521,** clearing the 0.50 benchmark from class notes. The clusters are no longer borderline arbitrary; the SPY-relative features produce groups that are meaningfully separated in feature space. 

Second, the cluster assignments are economically coherent in a way the previous version was not. **Utilities and Consumer Staples sit in the Defensive cluster 89% and 87% of the time respectively,** which is exactly what market fundamentals predict: these are low-beta, stable-cash-flow sectors that behave defensively regardless of the macro environment. Technology and Financials anchor the Cyclical cluster, consistent with their higher market sensitivity. 

Third, overall window-to-window stability is 77.3%, compared to the near-random switching observed in the correlation-only heatmap. The remaining instability is concentrated in sectors like Energy and Industrials that genuinely straddle the defensive-cyclical boundary depending on the macro regime, which is economically honest rather than a model failure.

------------------------------------------------------------------------

## 4. Kalman Filter

The Kalman Filter is a recursive algorithm that estimates the true underlying level of a noisy time series, updating its estimate each period as new data arrives. Applied to each cluster's cumulative return series, it separates genuine trend from daily noise. The result is an adaptive long-run level estimate that responds faster than a fixed 200-day moving average when conditions shift, but is not whipped around by single-day spikes. Because `dlmFilter()` runs a forward pass only, each day's estimate uses only data available up to that point. There is no look-ahead bias.

The model used is a local level model: the true level today equals the true level yesterday plus random drift. Two parameters control behaviour. `dV` is the observation variance, measuring how noisy the daily readings are. `dW` is the state evolution variance, measuring how fast the true level is allowed to change. A low `dW/dV` ratio produces a smoother, slower-moving filter. A high ratio lets the filter track the series more aggressively. The default ratio used here is 0.001, meaning the filter assumes the true level evolves slowly relative to daily noise. This parameter is tunable in the optimization section.

```{r kalman-setup}
#| message: false
#| warning: false
#| echo: false

# Identify Cyclical vs Defensive from static cluster assignments
cluster_beta_check <- feature_mat %>%
  dplyr::left_join(
    cluster_assignments_spy %>% dplyr::select(symbol, cluster),
    by = "symbol"
  ) %>%
  dplyr::group_by(cluster) %>%
  dplyr::summarise(mean_beta = mean(beta), .groups = "drop")

cyclical_id  <- cluster_beta_check %>% dplyr::slice_max(mean_beta, n = 1) %>% dplyr::pull(cluster)
defensive_id <- cluster_beta_check %>% dplyr::slice_min(mean_beta, n = 1) %>% dplyr::pull(cluster)

# Daily equal-weighted cluster returns using static assignments
cluster_daily <- ret_train %>%
  dplyr::filter(symbol %in% sector_tickers) %>%
  dplyr::left_join(
    cluster_assignments_spy %>% dplyr::select(symbol, cluster),
    by = "symbol"
  ) %>%
  dplyr::group_by(date, cluster) %>%
  dplyr::summarise(cluster_ret = mean(ret, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(
    cluster_label = dplyr::case_when(
      cluster == cyclical_id  ~ "Cyclical",
      cluster == defensive_id ~ "Defensive"
    )
  )

# Cumulative returns per cluster
cluster_cum <- cluster_daily %>%
  dplyr::arrange(cluster_label, date) %>%
  dplyr::group_by(cluster_label) %>%
  dplyr::mutate(cum_ret = cumsum(cluster_ret)) %>%
  dplyr::ungroup()

# Kalman Filter: local level model, forward pass only
KF_DW_SCALE <- 0.001  # dW/dV ratio -- tunable. Lower = slower KF = larger spread vs 20d SMA

apply_kf <- function(x, dw_scale = KF_DW_SCALE) {
  dv  <- var(diff(x), na.rm = TRUE)
  mod <- dlm::dlmModPoly(order = 1, dV = dv, dW = dv * dw_scale)
  flt <- dlm::dlmFilter(x, mod)
  as.numeric(dlm::dropFirst(flt$m))
}

# 20-day SMA
SMA_WINDOW <- 20

cluster_signals <- cluster_cum %>%
  dplyr::group_by(cluster_label) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    kf_level = apply_kf(cum_ret),
    sma_20   = TTR::SMA(cum_ret, n = SMA_WINDOW)
  ) %>%
  dplyr::ungroup()
```

```{r kalman-plot}
#| message: false
#| warning: false
#| echo: false

plot_df <- cluster_signals %>%
  dplyr::filter(!is.na(sma_20)) %>%
  tidyr::pivot_longer(
    cols      = c(cum_ret, kf_level, sma_20),
    names_to  = "series",
    values_to = "value"
  ) %>%
  dplyr::mutate(
    series = dplyr::recode(series,
      cum_ret  = "Cumulative Return",
      kf_level = "Kalman Level",
      sma_20   = "20d SMA"
    )
  )

colours <- c(
  "Cumulative Return" = "#888888",
  "Kalman Level"      = "#2c7bb6",
  "20d SMA"           = "#d7191c"
)

make_panel <- function(df, label, show_legend = TRUE) {
  d <- dplyr::filter(df, cluster_label == label)
  plotly::plot_ly(d, x = ~date, y = ~value, color = ~series,
                  colors = colours, type = "scatter", mode = "lines",
                  line = list(width = 1.2), showlegend = show_legend) %>%
    plotly::layout(
      annotations = list(list(
        text      = label, showarrow = FALSE,
        x         = 0.5, y = 1.02, xref = "paper", yref = "paper",
        font      = list(size = 13, color = "black"),
        xanchor   = "center"
      )),
      yaxis = list(title = "Cumulative Log Return"),
      xaxis = list(title = "")
    )
}

p_cyc <- make_panel(plot_df, "Cyclical", show_legend = TRUE)
p_def <- make_panel(plot_df, "Defensive", show_legend = FALSE)

plotly::subplot(p_cyc, p_def, nrows = 2, shareX = FALSE, titleY = TRUE) %>%
  plotly::layout(
    title  = list(
      text = paste0("Cluster Cumulative Returns: Raw vs Kalman Level vs 20d SMA<br>",
                    "<sup>Kalman dW/dV = ", KF_DW_SCALE,
                    ". 20d SMA = ", SMA_WINDOW, " days. Forward pass only. Zoom to inspect signal separation.</sup>"),
      x = 0.5
    ),
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.05)
  )
```

The grey line is the raw cumulative return for each cluster. The blue line is the Kalman-filtered level: the filter's best estimate of where the cluster truly sits after removing daily noise. The red line is the 20-day SMA. The momentum spread used in the signal is the vertical distance between the red and blue lines each day. When red is above blue, the cluster is running ahead of its long-run trend. When red is below blue, the cluster has pulled back. The stress state fires when both clusters show red below blue simultaneously.

------------------------------------------------------------------------

## 5. Signal Generation

### 5.1 Strategy Architecture

The complete daily decision process is summarized in the diagram below.

```{mermaid}
%%{init: {'theme': 'base', 'themeVariables': {'primaryColor': '#e0e0e0', 'primaryTextColor': '#0d1b2a', 'primaryBorderColor': '#555', 'lineColor': '#555', 'secondaryColor': '#fff2cc', 'tertiaryColor': '#ffffff'}}}%%
flowchart TD
    classDef startend fill:#cce5ff,stroke:#003366,stroke-width:1px;
    classDef process fill:#e0e0e0,stroke:#555,stroke-width:1px;
    classDef decision fill:#fff2cc,stroke:#777,stroke-width:1px;
    classDef stress fill:#fff2cc,stroke:#777,stroke-width:1px;
    classDef momentum fill:#ccffcc,stroke:#777,stroke-width:1px;
    classDef nosignal fill:#e0e0e0,stroke:#555,stroke-width:1px;

    A([Daily Close]):::startend
    B["Monthly Rebalance<br/>Cluster ETFs: Defensive / Cyclical<br/>via SPY-relative features"]:::process
    C["Daily: Equal-weighted cumulative<br/>return series per cluster"]:::process
    D["Kalman Filter per cluster<br/>Extract adaptive long-run level"]:::process
    E["20d SMA minus Kalman level<br/>= Momentum Spread per cluster"]:::process
    F{"Both clusters below<br/>Kalman level?<br/>(Market Stress)"}:::decision
    G["STRESS STATE<br/>10% Cyclical / 30% Defensive<br/>60% Cash"]:::stress
    H{"Cyclical spread minus<br/>Defensive spread<br/>> threshold?"}:::decision
    I["MOMENTUM CYCLICAL<br/>80% Cyclical / 20% Defensive"]:::momentum
    J{"Defensive spread minus<br/>Cyclical spread<br/>> threshold?"}:::decision
    K["MOMENTUM DEFENSIVE<br/>20% Cyclical / 80% Defensive"]:::stress
    L["NO SIGNAL<br/>Hold previous allocation"]:::nosignal
    M["Equal-weight ETFs<br/>within each cluster"]:::process
    N["Transaction costs: 0.05%<br/>per unit weight change"]:::process
    O([Execute at Open T+1]):::startend

    A --> B --> C --> D --> E --> F
    F -->|Yes| G
    F -->|No| H
    H -->|Yes| I
    H -->|No| J
    J -->|Yes| K
    J -->|No| L
    G --> M
    I --> M
    K --> M
    L --> M
    M --> N --> O
```

------------------------------------------------------------------------


### 5.2 Momentum Spread

The momentum spread is the 20d SMA minus the Kalman level for each cluster each day. A positive spread means the cluster is running above its long-run trend. A negative spread means it has pulled back below. The stress state fires when both spreads are negative simultaneously.

```{r spread-chart}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 6

spread_df <- cluster_signals %>%
  dplyr::filter(!is.na(sma_20)) %>%
  dplyr::mutate(spread = sma_20 - kf_level)

ggplot(spread_df, aes(x = date, y = spread)) +
  geom_line(aes(colour = cluster_label), linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "black", linewidth = 0.5) +
  geom_ribbon(aes(ymin = pmin(spread, 0), ymax = 0, fill = cluster_label),
              alpha = 0.15) +
  scale_colour_manual(values = c("Cyclical" = "#d7191c", "Defensive" = "#2c7bb6")) +
  scale_fill_manual(values   = c("Cyclical" = "#d7191c", "Defensive" = "#2c7bb6")) +
  facet_wrap(~ cluster_label, ncol = 1) +
  labs(
    title    = "Momentum Spread per Cluster: 20d SMA minus Kalman Level",
    subtitle = "Positive = cluster above long-run trend (momentum). Negative = below trend. Shaded area = below zero.",
    x        = NULL,
    y        = "Spread (cumulative log return units)",
    caption  = "Training period June 2000 to December 2024. Stress state fires when both clusters simultaneously negative."
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        strip.text      = element_text(face = "bold", size = 11))
```

```{r signal-export-build}
#| message: false
#| warning: false
#| echo: false
#| cache: true

# Cluster-level signals pivoted wide
signals_wide <- cluster_signals %>%
  dplyr::filter(!is.na(sma_20)) %>%
  dplyr::mutate(spread = sma_20 - kf_level) %>%
  tidyr::pivot_wider(
    id_cols     = date,
    names_from  = cluster_label,
    values_from = c(cum_ret, kf_level, sma_20, spread)
  ) %>%
  dplyr::rename(
    cyclical_cum_ret  = cum_ret_Cyclical,
    defensive_cum_ret = cum_ret_Defensive,
    cyclical_kf       = kf_level_Cyclical,
    defensive_kf      = kf_level_Defensive,
    cyclical_sma20    = sma_20_Cyclical,
    defensive_sma20   = sma_20_Defensive,
    cyclical_spread   = spread_Cyclical,
    defensive_spread  = spread_Defensive
  )

# Cluster label per ETF (static)
etf_clusters <- cluster_assignments_spy %>%
  dplyr::mutate(
    cluster_label = dplyr::if_else(cluster == cyclical_id, "Cyclical", "Defensive")
  ) %>%
  dplyr::select(symbol, cluster_label)

# Wide: one row per date, one column per ETF for ret and adjusted
ret_wide_export <- ret_train %>%
  dplyr::filter(symbol %in% sector_tickers) %>%
  dplyr::select(date, symbol, ret) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = ret,
                     names_prefix = "ret_")

adj_wide_export <- ret_train %>%
  dplyr::filter(symbol %in% sector_tickers) %>%
  dplyr::select(date, symbol, adjusted) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = adjusted,
                     names_prefix = "adj_")

# Cluster label row (static -- same every date, useful for Excel lookups)
# Forward-fill rolling cluster assignments from month-end to all trading dates
all_trading_dates <- ret_train %>%
  dplyr::filter(symbol == "XLK") %>%
  dplyr::arrange(date) %>%
  dplyr::select(date)

rolling_clusters_filled <- tidyr::expand_grid(
  date   = all_trading_dates$date,
  symbol = sector_tickers
) %>%
  dplyr::left_join(
    rolling_spy %>% dplyr::select(date, symbol, cluster),
    by = c("date", "symbol")
  ) %>%
  dplyr::arrange(symbol, date) %>%
  dplyr::group_by(symbol) %>%
  tidyr::fill(cluster, .direction = "down") %>%
  dplyr::ungroup() %>%
  dplyr::mutate(cluster = dplyr::if_else(cluster == "Cyclical", 1L, 0L))

# Pivot cluster labels wide: one column per ETF per date
cluster_wide_rolling <- rolling_clusters_filled %>%
  tidyr::pivot_wider(names_from = symbol, values_from = cluster,
                     names_prefix = "cluster_")

etf_export <- signals_wide %>%
  dplyr::left_join(ret_wide_export, by = "date") %>%
  dplyr::left_join(adj_wide_export, by = "date") %>%
  dplyr::left_join(cluster_wide_rolling, by = "date") %>%
  dplyr::arrange(date)
```

### 5.2 Signal Classification

```{r signal-classification}
#| message: false
#| warning: false
#| echo: false

SIGNAL_THRESHOLD    <- 0.010   # minimum absolute spread to trigger momentum signal
MAGNITUDE_THRESHOLD <- 0.080   # absolute spread level that triggers stress state
STRESS_DAYS         <- 3       # consecutive days stress must fire before confirming state

signal_states <- spread_df %>%
  dplyr::select(date, cluster_label, spread) %>%
  tidyr::pivot_wider(names_from = cluster_label, values_from = spread) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    # Raw stress flag each day
    stress_raw = (Cyclical < 0 & Defensive < 0) |
                 abs(Cyclical) > MAGNITUDE_THRESHOLD |
                 abs(Defensive) > MAGNITUDE_THRESHOLD,
    # Stress confirmed only after STRESS_DAYS consecutive days
    stress_confirmed = slider::slide_lgl(
      stress_raw, ~all(.x), .before = STRESS_DAYS - 1, .complete = TRUE
    ),
    stress_confirmed = tidyr::replace_na(stress_confirmed, FALSE),
    signal = dplyr::case_when(
      stress_confirmed                                          ~ "Stress",
      Cyclical > SIGNAL_THRESHOLD & Cyclical >= Defensive      ~ "Momentum Cyclical",
      Defensive > SIGNAL_THRESHOLD & Defensive > Cyclical      ~ "Momentum Defensive",
      TRUE                                                      ~ "No Signal"
    ),
    signal = factor(signal,
                    levels = c("Momentum Cyclical", "Momentum Defensive",
                               "Stress", "No Signal"))
  )
```

```{r signal-timeline}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 11
#| fig-height: 3

ggplot(signal_states, aes(x = date, y = 1, fill = signal)) +
  geom_tile() +
  scale_fill_manual(values = c(
    "Momentum Cyclical"  = "#d7191c",
    "Momentum Defensive" = "#2c7bb6",
    "Stress"             = "#f4a736",
    "No Signal"          = "#cccccc"
  )) +
  labs(
    title   = "Daily Signal State: Training Period",
    subtitle = paste0("Signal threshold = ", SIGNAL_THRESHOLD,
                      ". Magnitude threshold = ", MAGNITUDE_THRESHOLD, "."),
    x       = NULL,
    y       = NULL,
    fill    = "Signal State",
    caption = "Training period June 2000 to December 2024. Signal computed at Close(t), executed at Open(t+1)."
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y  = element_blank(),
        panel.grid   = element_blank(),
        legend.position = "bottom")
```

```{r signal-summary-table}
#| message: false
#| warning: false
#| echo: false

signal_states %>%
  dplyr::count(signal) %>%
  dplyr::mutate(pct = round(n / sum(n) * 100, 1)) %>%
  dplyr::rename(State = signal, Days = n, `%` = pct) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Signal State Frequency: Training Period",
    subtitle = "June 2000 to December 2024."
  ) %>%
  gt::fmt_number(columns = Days, decimals = 0) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::data_color(
    columns = `%`,
    palette = c("white", "#2c7bb6"),
    domain  = c(0, 60)
  )
```

------------------------------------------------------------------------

## 6. Execution and Implementation

### 6.1 Dividend Data

Before building the execution loop, dividend data is verified. Dividends are collected daily using ex-dividend dates from `quantmod::getDividends()`. On each ex-dividend date, cash increases by `shares_held * dividend_per_share`. This cash is included in portfolio value the following day, so it is proportionally reinvested through the normal weight calculation without any special reinvestment logic.

```{r dividend-pull}
#| message: false
#| warning: false
#| echo: false
#| cache: true

dividends_raw <- purrr::map_dfr(sector_tickers, ~ {
  x <- quantmod::getDividends(.x, from = "2000-06-01", to = "2026-03-31",
                               auto.assign = FALSE)
  if (is.null(x) || length(x) == 0) return(NULL)
  tibble::tibble(
    symbol   = .x,
    date     = zoo::index(x),
    dividend = as.numeric(x)
  )
}) %>%
  dplyr::filter(!is.na(dividend), dividend > 0) %>%
  dplyr::arrange(symbol, date)

# Summary table
dividends_raw %>%
  dplyr::group_by(symbol) %>%
  dplyr::summarise(
    `First Ex-Date`  = min(date),
    `Last Ex-Date`   = max(date),
    `Total Payments` = dplyr::n(),
    `Avg Dividend`   = round(mean(dividend), 4),
    `Total Paid`     = round(sum(dividend), 4),
    .groups          = "drop"
  ) %>%
  dplyr::rename(Symbol = symbol) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Dividend Data Verification: SPDR Sector ETFs",
    subtitle = "Source: quantmod::getDividends(). Ex-dividend dates and amounts."
  ) %>%
  gt::fmt_number(columns = c(`Avg Dividend`, `Total Paid`), decimals = 4) %>%
  gt::fmt_number(columns = `Total Payments`, decimals = 0) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_source_note("Dividends added to cash on ex-date. Reinvested proportionally through daily weight calculation.")
```

### 6.2 Data Preparation

The strategy is initialized with $1,000,000 in capital. A portfolio of this size is realistic for a high-net-worth individual or small institutional.

```{r execution-data-prep}
#| message: false
#| warning: false
#| echo: false

INITIAL_CAPITAL <- 1000000
TC_RATE         <- 0.0005   # 0.05% per unit of trade value

# Raw open and close prices (not adjusted -- dividends handled separately)
prices_wide <- raw %>%
  dplyr::filter(symbol %in% sector_tickers) %>%
  dplyr::select(symbol, date, open, close) %>%
  tidyr::pivot_wider(
    names_from  = symbol,
    values_from = c(open, close),
    names_glue  = "{symbol}_{.value}"
  ) %>%
  dplyr::arrange(date)

# SPY raw prices for benchmark
spy_prices <- raw %>%
  dplyr::filter(symbol == "SPY") %>%
  dplyr::select(date, spy_open = open, spy_close = close) %>%
  dplyr::arrange(date)

# Dividends wide: one column per ETF, 0 on non-dividend days
div_wide <- dividends_raw %>%
  dplyr::filter(date >= "2000-06-01", date <= "2026-03-31") %>%
  tidyr::pivot_wider(
    names_from   = symbol,
    values_from  = dividend,
    names_prefix = "div_",
    values_fill  = 0
  )

# Rolling cluster assignments forward-filled (1 = Cyclical, 0 = Defensive)
cluster_wide_daily <- rolling_clusters_filled %>%
  tidyr::pivot_wider(
    names_from  = symbol,
    values_from = cluster,
    names_prefix = "clust_"
  )

# Master loop dataframe: one row per trading day with all inputs
loop_data <- prices_wide %>%
  dplyr::left_join(div_wide,            by = "date") %>%
  dplyr::left_join(cluster_wide_daily,  by = "date") %>%
  dplyr::left_join(
    signal_states %>% dplyr::select(date, signal),
    by = "date"
  ) %>%
  dplyr::arrange(date) %>%
  # Training period only -- test period held out until Section 10
  dplyr::filter(!is.na(signal), date <= TRAIN_END) %>%
  # Replace NA dividends with 0
  dplyr::mutate(dplyr::across(
    dplyr::starts_with("div_"),
    ~tidyr::replace_na(.x, 0)
  )) %>%
  # Month-end flag for monthly cluster rebalancing
  dplyr::mutate(ym = lubridate::floor_date(date, "month")) %>%
  dplyr::group_by(ym) %>%
  dplyr::mutate(month_end_flag = date == max(date)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-ym)

# Sector allocation by signal state
get_alloc <- function(sig) {
  switch(as.character(sig),
    "Momentum Cyclical"  = list(cyc = 0.80, def = 0.20),
    "Momentum Defensive" = list(cyc = 0.20, def = 0.80),
    "Stress"             = list(cyc = 0.10, def = 0.30),
    list(cyc = 0.20, def = 0.80)   # fallback
  )
}
```

### 6.3 Execution Loop

The loop follows the same architecture as the dynamic portfolio rebalancing exercise: history matrices pre-allocated, state carried forward day-to-day, signal lag enforced by separating signal generation (Close t) from execution (Open t+1). Portfolio value is tracked as `sum(shares * close_prices) + cash`, which correctly captures close-to-open returns on sold positions, open-to-close returns on purchased positions, and close-to-close returns on unchanged positions.

```{r execution-loop}
#| message: false
#| warning: false
#| echo: false
#| cache: true

n_days <- nrow(loop_data)

# History matrices -- pre-allocate for performance
portfolio_val_hist  <- numeric(n_days)
cash_hist           <- numeric(n_days)
tc_hist             <- numeric(n_days)
signal_hist         <- character(n_days)
shares_hist         <- matrix(0, nrow = n_days, ncol = length(sector_tickers),
                               dimnames = list(NULL, sector_tickers))
weight_hist         <- matrix(0, nrow = n_days, ncol = length(sector_tickers),
                               dimnames = list(NULL, sector_tickers))

# State variables
current_cash        <- INITIAL_CAPITAL
current_shares      <- setNames(rep(0, length(sector_tickers)), sector_tickers)
pending_signal      <- NA_character_   # signal set at Close(t), executed at Open(t+1)
last_active_signal  <- NA_character_   # last non-No-Signal state executed
prev_pv             <- INITIAL_CAPITAL # portfolio value at previous close

for (i in seq_len(n_days)) {

  today        <- loop_data[i, ]
  today_signal <- as.character(today$signal)

  # ── STEP 1: Collect dividends at start of day ──────────────────────────────
  div_income <- sum(sapply(sector_tickers, function(etf) {
    div_col <- paste0("div_", etf)
    if (div_col %in% names(today)) {
      dv <- today[[div_col]]
      if (!is.na(dv) && dv > 0) return(current_shares[[etf]] * dv)
    }
    0
  }), na.rm = TRUE)

  current_cash <- current_cash + div_income

  # ── STEP 2: Execute pending order at Open(t) ───────────────────────────────
  trade_cost <- 0

  # Rebalance on signal change OR on month-end (captures cluster membership shifts)
  is_month_end <- isTRUE(loop_data$month_end_flag[i])
  exec_sig <- if (!is.na(pending_signal) && pending_signal != "No Signal") {
    pending_signal
  } else if (is_month_end && !is.na(last_active_signal) && last_active_signal != "No Signal") {
    last_active_signal
  } else {
    NA_character_
  }

  if (!is.na(exec_sig)) {
    alloc <- get_alloc(exec_sig)

    # Count available ETFs per cluster today (non-NA open price)
    clust_today <- setNames(sapply(sector_tickers, function(etf) {
      col <- paste0("clust_", etf); if (col %in% names(today)) today[[col]] else NA_integer_
    }), sector_tickers)

    avail_cyc <- sector_tickers[!is.na(clust_today) & clust_today == 1 &
                                  !is.na(sapply(sector_tickers, function(e) today[[paste0(e, "_open")]]))]
    avail_def <- sector_tickers[!is.na(clust_today) & clust_today == 0 &
                                  !is.na(sapply(sector_tickers, function(e) today[[paste0(e, "_open")]]))]

    n_cyc <- length(avail_cyc)
    n_def <- length(avail_def)

    # Target weight per ETF: cluster allocation / count of available ETFs in cluster
    target_w <- setNames(rep(0, length(sector_tickers)), sector_tickers)
    if (n_cyc > 0) target_w[avail_cyc] <- alloc$cyc / n_cyc
    if (n_def > 0) target_w[avail_def] <- alloc$def / n_def

    # Target dollars based on previous close portfolio value
    target_dollars <- prev_pv * target_w

    # Execute at open: compute new shares and cash flows
    new_shares <- current_shares
    for (etf in sector_tickers) {
      open_p <- today[[paste0(etf, "_open")]]
      if (is.na(open_p) || open_p <= 0) next
      new_shares[[etf]]  <- target_dollars[[etf]] / open_p
      delta_shares       <- new_shares[[etf]] - current_shares[[etf]]
      current_cash       <- current_cash - (delta_shares * open_p)
      trade_cost         <- trade_cost + abs(delta_shares * open_p) * TC_RATE
    }

    current_cash   <- current_cash - trade_cost
    current_shares <- new_shares
    if (!is.na(pending_signal) && pending_signal != "No Signal")
      last_active_signal <- pending_signal
  }

  # ── STEP 3: Mark portfolio to close ────────────────────────────────────────
  equity_value <- sum(sapply(sector_tickers, function(etf) {
    close_p <- today[[paste0(etf, "_close")]]
    if (!is.na(close_p) && close_p > 0) current_shares[[etf]] * close_p else 0
  }), na.rm = TRUE)

  portfolio_value <- equity_value + current_cash

  # ── STEP 4: Store history ───────────────────────────────────────────────────
  portfolio_val_hist[i] <- portfolio_value
  cash_hist[i]          <- current_cash
  tc_hist[i]            <- trade_cost
  signal_hist[i]        <- today_signal
  shares_hist[i, ]      <- current_shares

  if (portfolio_value > 0) {
    for (etf in sector_tickers) {
      close_p <- today[[paste0(etf, "_close")]]
      if (!is.na(close_p) && close_p > 0)
        weight_hist[i, etf] <- (current_shares[[etf]] * close_p) / portfolio_value
    }
  }

  # ── STEP 5: Set pending signal for tomorrow ─────────────────────────────────
  # Pending fires when signal changes AND new signal is active (not No Signal)
  prev_signal    <- if (i == 1) NA_character_ else signal_hist[i - 1]
  pending_signal <- dplyr::case_when(
    is.na(prev_signal) & today_signal != "No Signal" ~ today_signal,  # day 1 entry
    !is.na(prev_signal) & today_signal != prev_signal &
      today_signal != "No Signal"                    ~ today_signal,  # signal change
    TRUE                                             ~ NA_character_   # no trade
  )

  prev_pv <- portfolio_value
}

# Assemble results dataframe
backtest_results <- loop_data %>%
  dplyr::select(date) %>%
  dplyr::mutate(
    portfolio_value = portfolio_val_hist,
    cash            = cash_hist,
    tc              = tc_hist,
    signal          = signal_hist
  ) %>%
  dplyr::bind_cols(
    as_tibble(weight_hist) %>%
      dplyr::rename_with(~paste0("w_", .x))
  )
```

### 6.4 Implementation Assumptions

**Signal timing and execution lag.** The signal is generated at each day's close using only information available up to that point. No position change is executed until the following day's open. This one-day lag is enforced throughout the loop by storing a pending signal at the end of each day and acting on it at the start of the next. The practical implication is that the strategy can never trade on the same day it detects a signal, which eliminates a common source of look-ahead bias in backtesting.

**Dollar weighting at previous close.** Target dollar allocations are computed using the portfolio value at the previous day's close, and shares are determined by dividing those dollar targets by the next morning's open price. This means the capital deployed per ETF is fixed the night before, and the open price only determines share count. The result is that there is no slippage from execution order: sells and buys net to zero regardless of which ETF is processed first. The overnight gap between yesterday's close and today's open is a real return that is captured in the portfolio value path, not a source of error.

**Partial shares.** The loop assumes fractional share ownership, so a \$16,400 allocation to XLK at \$230 per share results in 71.30 shares held. Real ETF execution requires whole shares, which introduces a small rounding residual at each rebalance. At a \$1,000,000 portfolio level this error is economically negligible, and the assumption is standard practice in academic backtesting. It is acknowledged as a deviation from live execution.

**Dividends.** Dividends are collected on the ex-dividend date using `quantmod::getDividends()` and added directly to cash. The following day, when portfolio value is used to compute target dollar allocations, the dividend cash is included proportionally across all positions through the normal weight calculation. This is equivalent to immediate reinvestment without requiring a separate reinvestment step. The timing of dividends relative to ex-date versus payment date introduces a small lag versus real cash flows, but ex-date is the correct economic event for a mark-to-market portfolio.

**Transaction costs.** A flat rate of 0.05% is applied to the absolute value of each trade on execution days, including the initial entry. This is a conservative estimate for liquid large-cap ETFs, where bid-ask spreads and market impact are typically lower. The rate is applied symmetrically to both buys and sells. No financing costs are charged on the cash position held during stress periods, which slightly understates the cost of capital preservation relative to a real fund that would earn the risk-free rate on that cash.

**No market impact.** The backtest assumes all trades execute at the published open price regardless of size. A \$1,000,000 portfolio rotating 60% of capital in a single session is unlikely to move prices on SPY sector ETFs, which trade billions of dollars daily. The assumption is reasonable at this portfolio scale and would require revisiting if the strategy were scaled to institutional size.

**Cash earns zero.** The 60% cash allocation in the stress state earns no return. In practice this capital would sit in a money market fund or short-duration Treasuries earning the risk-free rate, which over a multi-month stress period is a meaningful omission. This assumption is conservative: it understates strategy returns during stress periods and therefore does not flatter the results.

------------------------------------------------------------------------

## 7. Training Period Backtest (Pre-Optimization)

```{r backtest-prep}
#| message: false
#| warning: false
#| echo: false

# Strategy daily returns
strategy_returns <- backtest_results %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    strat_ret = portfolio_value / dplyr::lag(portfolio_value) - 1
  ) %>%
  dplyr::filter(!is.na(strat_ret))

# SPY benchmark: adjusted close (dividends already embedded)
spy_bench <- raw %>%
  dplyr::filter(symbol == "SPY", date >= min(strategy_returns$date),
                date <= TRAIN_END) %>%
  dplyr::select(date, spy_adj = adjusted) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    spy_ret   = spy_adj / dplyr::lag(spy_adj) - 1,
    spy_value = 1000000 * (spy_adj / dplyr::first(spy_adj))
  )

# Combined for plotting
combined <- strategy_returns %>%
  dplyr::left_join(spy_bench %>% dplyr::select(date, spy_ret, spy_value),
                   by = "date") %>%
  dplyr::mutate(
    strat_idx = portfolio_value / dplyr::first(portfolio_value) * 100,
    spy_idx   = spy_value / dplyr::first(spy_value) * 100,
    # Trade flag: 1 if signal changed and trade fired
    trade_flag = dplyr::if_else(tc > 0, 1L, 0L),
    # Position encoding
    pos_num = dplyr::case_when(
      signal == "Momentum Cyclical"  ~  1,
      signal == "No Signal"          ~  0,
      signal == "Momentum Defensive" ~ -1,
      signal == "Stress"             ~ -2,
      TRUE                           ~  0
    ),
    # Drawdowns
    strat_peak = cummax(portfolio_value),
    spy_peak   = cummax(spy_value),
    strat_dd   = (portfolio_value - strat_peak) / strat_peak,
    spy_dd     = (spy_value - spy_peak) / spy_peak
  )

# Helper: annualised return
ann_ret <- function(r) {
  n_years <- length(r) / 252
  prod(1 + r, na.rm = TRUE)^(1 / n_years) - 1
}

# Helper: max drawdown
max_dd <- function(r) {
  cum <- cumprod(1 + r)
  peak <- cummax(cum)
  min((cum - peak) / peak, na.rm = TRUE)
}

# Helper: Omega ratio (threshold = 0)
omega_ratio <- function(r, threshold = 0) {
  gains  <- sum(pmax(r - threshold, 0), na.rm = TRUE)
  losses <- sum(pmax(threshold - r, 0), na.rm = TRUE)
  if (losses == 0) return(NA_real_)
  gains / losses
}

# Helper: Calmar
calmar <- function(r) {
  ann <- ann_ret(r)
  mdd <- abs(max_dd(r))
  if (mdd == 0) return(NA_real_)
  ann / mdd
}

strat_r <- strategy_returns$strat_ret
spy_r   <- spy_bench$spy_ret[!is.na(spy_bench$spy_ret)]
```

### 7.1 Strategy Results Chart

```{r xts-chart}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-keep: last

# Convert to xts for quantmod-style multi-panel chart
tick <- combined %>%
  dplyr::select(date, strat_idx, trade_flag, pos_num, strat_idx) %>%
  timetk::tk_xts(date_var = date)

plot(tick$strat_idx,
     main       = "Strategy Results",
     ylab       = "Indexed to 100",
     col        = "black",
     lwd        = 1.2,
     grid.ticks.on = "years")

xts::addSeries(
  tick$trade_flag,
  main = "Trades",
  on   = NA,
  type = "h",
  col  = "#2c7bb6",
  lwd  = 1,
  pch  = 0
)

xts::addSeries(
  tick$pos_num,
  main = "Position (1=Cyc, -1=Def, -2=Stress)",
  on   = NA,
  type = "h",
  col  = "#2c7bb6",
  lwd  = 1,
  pch  = 0
)

xts::addSeries(
  tick$strat_idx,
  main = "CumEQ",
  on   = NA,
  type = "l",
  col  = "#2c7bb6",
  lwd  = 1,
  pch  = 0
)
```

### 7.2 Strategy vs SPY Equity Curve

```{r equity-curve}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 5

combined %>%
  dplyr::select(date, Strategy = strat_idx, SPY = spy_idx) %>%
  tidyr::pivot_longer(-date, names_to = "series", values_to = "value") %>%
  ggplot(aes(x = date, y = value, colour = series, linewidth = series)) +
  geom_line(alpha = 0.9) +
  scale_colour_manual(values = c("Strategy" = "#2c7bb6", "SPY" = "#d7191c")) +
  scale_linewidth_manual(values = c("Strategy" = 0.9, "SPY" = 0.6)) +
  labs(
    title    = "Strategy vs SPY: Cumulative Return (Indexed to 100)",
    subtitle = "Training period. SPY uses adjusted close with dividends reinvested.",
    x        = NULL, y = "Index (100 = start)", colour = NULL, linewidth = NULL,
    caption  = "Training period June 2000 to December 2024."
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

### 7.3 Drawdown Comparison

```{r drawdown-chart}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 4

combined %>%
  dplyr::select(date, Strategy = strat_dd, SPY = spy_dd) %>%
  tidyr::pivot_longer(-date, names_to = "series", values_to = "dd") %>%
  ggplot(aes(x = date, y = dd * 100, colour = series, fill = series)) +
  geom_ribbon(aes(ymin = dd * 100, ymax = 0), alpha = 0.2) +
  geom_line(linewidth = 0.6) +
  scale_colour_manual(values = c("Strategy" = "#2c7bb6", "SPY" = "#d7191c")) +
  scale_fill_manual(values   = c("Strategy" = "#2c7bb6", "SPY" = "#d7191c")) +
  labs(
    title    = "Drawdown: Strategy vs SPY",
    subtitle = "Drawdown from rolling peak. Strategy stress state targets capital preservation during severe drawdowns.",
    x        = NULL, y = "Drawdown (%)", colour = NULL, fill = NULL,
    caption  = "Training period June 2000 to December 2024."
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

### 7.4 Performance Metrics

```{r performance-table}
#| message: false
#| warning: false
#| echo: false

tibble::tibble(
  Metric = c("Total Return (%)", "Annualised Return (%)", "Sharpe Ratio",
             "Omega Ratio", "Sortino Ratio", "Max Drawdown (%)",
             "Calmar Ratio", "Hit Rate (%)"),
  Strategy = c(
    round((dplyr::last(combined$portfolio_value) / 1e6 - 1) * 100, 2),
    round(ann_ret(strat_r) * 100, 2),
    round(mean(strat_r, na.rm = TRUE) / sd(strat_r, na.rm = TRUE) * sqrt(252), 3),
    round(omega_ratio(strat_r), 3),
    round(mean(strat_r[strat_r > 0], na.rm = TRUE) /
            abs(mean(strat_r[strat_r < 0], na.rm = TRUE)) *
            sqrt(252), 3),
    round(max_dd(strat_r) * 100, 2),
    round(calmar(strat_r), 3),
    round(mean(strat_r > 0, na.rm = TRUE) * 100, 2)
  ),
  SPY = c(
    round((dplyr::last(combined$spy_value) / 1e6 - 1) * 100, 2),
    round(ann_ret(spy_r) * 100, 2),
    round(mean(spy_r, na.rm = TRUE) / sd(spy_r, na.rm = TRUE) * sqrt(252), 3),
    round(omega_ratio(spy_r), 3),
    round(mean(spy_r[spy_r > 0], na.rm = TRUE) /
            abs(mean(spy_r[spy_r < 0], na.rm = TRUE)) *
            sqrt(252), 3),
    round(max_dd(spy_r) * 100, 2),
    round(calmar(spy_r), 3),
    round(mean(spy_r > 0, na.rm = TRUE) * 100, 2)
  )
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Training Period Performance: Strategy vs SPY",
    subtitle = "June 2000 to December 2024. Strategy initialized at $1,000,000."
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#f0f4ff"),
    locations = gt::cells_body(columns = Strategy)
  ) %>%
  gt::tab_source_note("Sharpe and Sortino annualised using sqrt(252). Omega threshold = 0. Calmar = annualised return / abs(max drawdown).")
```

------------------------------------------------------------------------

## 8. Risk Appetite

### 8.1 Maximum Drawdown Tolerance

The strategy targets a maximum drawdown below 40%. This threshold is set with reference to the training period result of 35.1%, which demonstrates the stress state is doing meaningful work relative to SPY's 55.2% drawdown over the same period. A 40% hard cap is deliberately conservative relative to the training result, leaving headroom for the testing period where out-of-sample deterioration is expected. Most institutional mandates for multi-asset equity strategies sit in the 20-30% drawdown tolerance range. This strategy does not claim to meet that bar at current parameters, and the optimization section will explicitly test whether tighter stress triggers can push the drawdown lower without unacceptable return sacrifice.

### 8.2 Optimization Objective

The primary optimization objective is the **Omega ratio:** the ratio of cumulative gains above zero to cumulative losses below zero across all daily returns. Unlike Sharpe, which summarizes the distribution with only mean and variance, Omega uses the full return distribution and treats upside and downside asymmetrically. A strategy with Omega above 1.0 generates more total gains than losses; the higher the ratio, the more favorably the return distribution is skewed toward the investor. This is the right objective for a sector rotation strategy where the goal is to avoid large left-tail events rather than simply maximize average return. Calmar ratio is used as a secondary filter: only parameter combinations with a Calmar above 0.10 are retained, which eliminates cases where Omega looks acceptable but the strategy is hiding an unacceptable peak-to-trough decline. A pure return maximization objective would push the optimizer toward parameters that keep the strategy fully invested at all times, which would defeat the purpose of the stress state.

### 8.3 Position Sizing

Position sizing is equal-weight within each cluster, with cluster-level weights determined by the active signal state. The equal-weight convention is a deliberate choice rather than a constraint. It avoids concentration in whichever sector has the highest recent momentum, distributes risk across all ETFs within each cluster, and ensures the strategy is testing the cluster signal rather than a size-tilt signal. The known cost is underperformance relative to a cap-weighted benchmark during periods when a small number of mega-cap sectors dominate market returns. This is accepted as a structural feature of the approach.

The four signal states and their corresponding allocations are:

```{r risk-appetite-table}
#| message: false
#| warning: false
#| echo: false

tibble::tibble(
  `Signal State`       = c("Momentum Cyclical", "Momentum Defensive",
                            "Stress", "No Signal"),
  `Cyclical Allocation` = c("80%", "20%", "10%", "Hold previous"),
  `Defensive Allocation` = c("20%", "80%", "30%", "Hold previous"),
  `Cash`               = c("0%", "0%", "60%", "Hold previous"),
  `Trigger`            = c(
    "Cyclical spread > signal threshold AND Cyclical >= Defensive",
    "Defensive spread > signal threshold AND Defensive > Cyclical",
    "Both spreads negative OR either spread exceeds magnitude threshold for 3 consecutive days",
    "Spread difference below threshold: no trade, hold last active allocation"
  )
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Signal States and Target Allocations",
    subtitle = "Equal-weight within each cluster. Allocations applied to portfolio value at previous close."
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::cols_width(
    `Signal State`         ~ px(160),
    `Cyclical Allocation`  ~ px(120),
    `Defensive Allocation` ~ px(130),
    Cash                   ~ px(80),
    Trigger                ~ px(300)
  ) %>%
  gt::tab_source_note("Per-ETF weight = cluster allocation / number of available ETFs in cluster on that date.")
```

### 8.4 Tunable Parameters and Optimization Grid

Four parameters govern the strategy's risk posture. All four are candidates for the optimization grid search in Section 9.

```{r params-table}
#| message: false
#| warning: false
#| echo: false

tibble::tibble(
  Parameter             = c("Signal Threshold", "Magnitude Threshold",
                             "Stress Confirmation Days", "SMA Window"),
  Variable              = c("SIGNAL_THRESHOLD", "MAGNITUDE_THRESHOLD",
                             "STRESS_DAYS", "SMA_WINDOW"),
  Default               = c("0.010", "0.080", "3", "20"),
  `Optimization Range`  = c("0.005 to 0.030", "0.050 to 0.120",
                             "2 to 5", "10 to 30"),
  `Risk Interpretation` = c(
    "Higher = fewer momentum signals, less trading, more time in No Signal",
    "Higher = stress state fires less often, more equity exposure during volatile periods",
    "Higher = stress confirmation slower, reduces whipsawing into cash",
    "Shorter = faster momentum signal, more responsive but noisier"
  )
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Tunable Parameters",
    subtitle = "All parameters fitted on training data only. Applied unchanged to testing period."
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::cols_width(
    Parameter            ~ px(160),
    Variable             ~ px(160),
    Default              ~ px(80),
    `Optimization Range` ~ px(140),
    `Risk Interpretation`~ px(280)
  ) %>%
  gt::tab_source_note("KF dW/dV ratio fixed at 0.001. Lookback window fixed at 60 days. Both candidates for extended optimization given time.")
```

------------------------------------------------------------------------

## 9. Optimization

The strategy has four tunable parameters: the rolling lookback window for clustering features, the SMA window used to compute the momentum spread, the absolute spread threshold that triggers a directional signal, and the magnitude threshold that triggers the stress state. Rather than hand-picking these by inspection (which would introduce selection bias), a grid search is run on the training period only. The objective is Omega Ratio: it captures the full return distribution above and below a threshold, not just the ratio of mean return to volatility. Calmar Ratio is used as a secondary filter to ensure winning parameter combinations are not hiding an unacceptable drawdown profile.

Two optimization surfaces are produced. The first varies lookback window and SMA window while holding thresholds fixed. The second varies signal and magnitude thresholds while holding the window parameters fixed. This decomposition reflects the natural structure of the strategy: the window parameters govern how the signal is constructed, and the threshold parameters govern when it fires.

### 9.1 Strategy Function

```{r opt-setup}
#| message: false
#| warning: false
#| echo: false

run_strategy <- function(lookback_roll, sma_window, sig_thresh, mag_thresh,
                         excl_start = NULL, excl_end = NULL) {

  # ── Rolling cluster assignments ──────────────────────────────────────────────
  all_d <- sector_train %>%
    dplyr::filter(symbol == "XLK") %>%
    dplyr::arrange(date) %>%
    dplyr::pull(date)

  me <- sector_train %>%
    dplyr::filter(symbol == "XLK") %>%
    dplyr::mutate(ym = lubridate::floor_date(date, "month")) %>%
    dplyr::group_by(ym) %>%
    dplyr::slice_max(date, n = 1) %>%
    dplyr::ungroup() %>%
    dplyr::filter(date >= all_d[lookback_roll]) %>%
    dplyr::pull(date)

  cw <- function(end_date) {
    ws   <- sector_train %>%
      dplyr::filter(date <= end_date) %>%
      dplyr::group_by(symbol) %>%
      dplyr::slice_tail(n = lookback_roll) %>%
      dplyr::ungroup()
    wspy <- spy_train %>%
      dplyr::filter(date <= end_date) %>%
      dplyr::slice_tail(n = lookback_roll)

    sc  <- sum(wspy$ret, na.rm = TRUE)
    sv  <- var(wspy$ret, na.rm = TRUE)
    smd <- max_dd(wspy$ret)

    feats <- ws %>%
      dplyr::group_by(symbol) %>%
      dplyr::summarise(
        excess_ret = sum(ret, na.rm = TRUE) - sc,
        beta       = cov(ret, wspy$ret[seq_len(dplyr::n())],
                         use = "complete.obs") / sv,
        mdd_ratio  = max_dd(ret) / smd,
        .groups    = "drop"
      )

    mat <- scale(feats %>% dplyr::select(excess_ret, beta, mdd_ratio))
    rownames(mat) <- feats$symbol
    labs <- cutree(hclust(dist(mat, method = "euclidean"), method = "ward.D2"), k = 2)
    mb   <- tapply(feats$beta, labs, mean)
    cr   <- as.integer(names(which.max(mb)))
    tibble::tibble(date    = end_date,
                   symbol  = feats$symbol,
                   cluster = ifelse(labs == cr, "Cyclical", "Defensive"))
  }

  rolling_r <- purrr::map_dfr(me, cw)

  atd <- sector_train %>%
    dplyr::filter(symbol == "XLK") %>%
    dplyr::arrange(date) %>%
    dplyr::select(date)

  rf <- tidyr::expand_grid(date = atd$date, symbol = sector_tickers) %>%
    dplyr::left_join(rolling_r %>% dplyr::select(date, symbol, cluster),
                     by = c("date", "symbol")) %>%
    dplyr::arrange(symbol, date) %>%
    dplyr::group_by(symbol) %>%
    tidyr::fill(cluster, .direction = "down") %>%
    dplyr::ungroup() %>%
    dplyr::mutate(cluster = dplyr::if_else(cluster == "Cyclical", 1L, 0L))

  cwd <- rf %>%
    tidyr::pivot_wider(names_from  = symbol,
                       values_from = cluster,
                       names_prefix = "clust_")

  # ── KF + SMA + spread ────────────────────────────────────────────────────────
  cs <- cluster_daily %>%
    dplyr::group_by(cluster_label) %>%
    dplyr::arrange(date) %>%
    dplyr::mutate(
      cum_ret  = cumsum(cluster_ret),
      kf_level = apply_kf(cum_ret, dw_scale = KF_DW_SCALE),
      sma_n    = TTR::SMA(cum_ret, n = sma_window)
    ) %>%
    dplyr::ungroup()

  sr <- cs %>%
    dplyr::filter(!is.na(sma_n)) %>%
    dplyr::mutate(spread = sma_n - kf_level) %>%
    dplyr::select(date, cluster_label, spread) %>%
    tidyr::pivot_wider(names_from = cluster_label, values_from = spread) %>%
    dplyr::arrange(date) %>%
    dplyr::mutate(
      stress_raw = (Cyclical < 0 & Defensive < 0) |
                   abs(Cyclical) > mag_thresh      |
                   abs(Defensive) > mag_thresh,
      stress_confirmed = slider::slide_lgl(
        stress_raw, ~all(.x), .before = STRESS_DAYS - 1, .complete = TRUE
      ),
      stress_confirmed = tidyr::replace_na(stress_confirmed, FALSE),
      signal = dplyr::case_when(
        stress_confirmed                                  ~ "Stress",
        Cyclical  > sig_thresh & Cyclical  >= Defensive  ~ "Momentum Cyclical",
        Defensive > sig_thresh & Defensive >  Cyclical   ~ "Momentum Defensive",
        TRUE                                              ~ "No Signal"
      )
    )

  # ── Loop data ─────────────────────────────────────────────────────────────────
  ld <- prices_wide %>%
    dplyr::left_join(div_wide, by = "date") %>%
    dplyr::left_join(cwd,      by = "date") %>%
    dplyr::left_join(sr %>% dplyr::select(date, signal), by = "date") %>%
    dplyr::arrange(date) %>%
    dplyr::filter(!is.na(signal), date <= TRAIN_END) %>%
    dplyr::mutate(dplyr::across(dplyr::starts_with("div_"),
                                ~tidyr::replace_na(.x, 0)))

  # Optional: exclude a stress period from performance measurement
  if (!is.null(excl_start) && !is.null(excl_end)) {
    ld <- ld %>% dplyr::filter(date < as.Date(excl_start) | date > as.Date(excl_end))
  }

  n     <- nrow(ld)
  pv_h  <- numeric(n)
  sig_h <- character(n)

  cur_cash   <- INITIAL_CAPITAL
  cur_shares <- setNames(rep(0, length(sector_tickers)), sector_tickers)
  pend_sig   <- NA_character_
  prev_pv    <- INITIAL_CAPITAL

  for (i in seq_len(n)) {
    td  <- ld[i, ]
    tds <- as.character(td$signal)

    div_inc <- sum(sapply(sector_tickers, function(etf) {
      dc <- paste0("div_", etf)
      if (dc %in% names(td)) {
        dv <- td[[dc]]
        if (!is.na(dv) && dv > 0) return(cur_shares[[etf]] * dv)
      }
      0
    }), na.rm = TRUE)
    cur_cash <- cur_cash + div_inc

    tc <- 0
    if (!is.na(pend_sig) && pend_sig != "No Signal") {
      alloc <- get_alloc(pend_sig)
      ct <- setNames(sapply(sector_tickers, function(etf) {
        col <- paste0("clust_", etf)
        if (col %in% names(td)) td[[col]] else NA_integer_
      }), sector_tickers)
      ac <- sector_tickers[!is.na(ct) & ct == 1 &
              !is.na(sapply(sector_tickers, function(e) td[[paste0(e, "_open")]]))]
      ad <- sector_tickers[!is.na(ct) & ct == 0 &
              !is.na(sapply(sector_tickers, function(e) td[[paste0(e, "_open")]]))]
      tw <- setNames(rep(0, length(sector_tickers)), sector_tickers)
      if (length(ac) > 0) tw[ac] <- alloc$cyc / length(ac)
      if (length(ad) > 0) tw[ad] <- alloc$def / length(ad)
      tdols <- prev_pv * tw
      ns    <- cur_shares
      for (etf in sector_tickers) {
        op <- td[[paste0(etf, "_open")]]
        if (is.na(op) || op <= 0) next
        ns[[etf]]  <- tdols[[etf]] / op
        dlt        <- ns[[etf]] - cur_shares[[etf]]
        cur_cash   <- cur_cash - dlt * op
        tc         <- tc + abs(dlt * op) * TC_RATE
      }
      cur_cash   <- cur_cash - tc
      cur_shares <- ns
    }

    eq <- sum(sapply(sector_tickers, function(etf) {
      cp <- td[[paste0(etf, "_close")]]
      if (!is.na(cp) && cp > 0) cur_shares[[etf]] * cp else 0
    }), na.rm = TRUE)

    pv        <- eq + cur_cash
    pv_h[i]   <- pv
    sig_h[i]  <- tds

    ps       <- if (i == 1) NA_character_ else sig_h[i - 1]
    pend_sig <- dplyr::case_when(
      is.na(ps)  & tds != "No Signal"              ~ tds,
      !is.na(ps) & tds != ps & tds != "No Signal"  ~ tds,
      TRUE                                          ~ NA_character_
    )
    prev_pv <- pv
  }

  r <- diff(pv_h) / head(pv_h, -1)
  r <- r[is.finite(r)]

  sortino_r <- {
    excess <- r - 0
    dd_sq  <- mean(pmin(excess, 0)^2, na.rm = TRUE)
    if (dd_sq == 0) NA_real_ else mean(excess, na.rm = TRUE) * 252 / sqrt(dd_sq * 252)
  }

  tibble::tibble(
    lookback_roll = lookback_roll,
    sma_window    = sma_window,
    sig_thresh    = sig_thresh,
    mag_thresh    = mag_thresh,
    omega         = omega_ratio(r),
    calmar_r      = calmar(r),
    sharpe        = mean(r, na.rm = TRUE) / sd(r, na.rm = TRUE) * sqrt(252),
    sortino       = sortino_r,
    ann_ret_pct   = ann_ret(r) * 100,
    max_dd_pct    = round(max_dd(r) * 100, 2),
    hit_rate      = mean(r > 0, na.rm = TRUE) * 100
  )
}
```

### 9.2 Grid Search

Two 10-by-10 grids are searched in parallel using `doParallel`. Thresholds are optimized first because they have the most direct effect on strategy behaviour: they determine how often the strategy enters a position at all. Grid 1 varies the signal threshold (0.003 to 0.030) and the magnitude threshold (0.010 to 0.100), with window parameters held at their defaults. The magnitude threshold range is extended down to 0.010 to capture the region where the surface showed the most sensitivity. Grid 2 then varies the lookback window and SMA window using the best thresholds identified in Grid 1. Results are cached to avoid re-running the full search on every render.

```{r opt-grid1}
#| message: false
#| warning: false
#| echo: false
#| cache: true

grid1 <- tidyr::expand_grid(
  sig_thresh = c(0.005, 0.010, 0.015, 0.020, 0.025, 0.030, 0.040, 0.050, 0.060, 0.080),
  mag_thresh = c(0.010, 0.020, 0.030, 0.040, 0.050, 0.060, 0.070, 0.080, 0.090, 0.100)
)

n_cores <- max(1, parallel::detectCores() - 1)
cl1 <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl1)

parallel::clusterExport(cl1, varlist = c(
  "run_strategy",
  "sector_train", "spy_train", "prices_wide", "div_wide",
  "cluster_daily", "sector_tickers",
  "max_dd", "apply_kf", "get_alloc", "omega_ratio", "calmar", "ann_ret",
  "TRAIN_END", "STRESS_DAYS", "KF_DW_SCALE", "INITIAL_CAPITAL", "TC_RATE"
), envir = globalenv())

invisible(parallel::clusterEvalQ(cl1, {
  library(tidyverse); library(dlm); library(TTR); library(slider); library(lubridate)
}))

opt1 <- foreach::foreach(
  i         = seq_len(nrow(grid1)),
  .combine  = dplyr::bind_rows,
  .packages = c("dplyr", "tidyr", "purrr", "lubridate", "TTR", "dlm", "slider")
) %dopar% {
  run_strategy(
    lookback_roll = 60,
    sma_window    = 20,
    sig_thresh    = grid1$sig_thresh[i],
    mag_thresh    = grid1$mag_thresh[i]
  )
}

parallel::stopCluster(cl1)
```

```{r opt-grid2}
#| message: false
#| warning: false
#| echo: false
#| cache: true

# Best thresholds from Grid 1: max Omega subject to Calmar > 0.10
best1 <- opt1 %>%
  dplyr::filter(calmar_r > 0.10) %>%
  dplyr::slice_max(omega, n = 1)

best_sig <- if (nrow(best1) > 0) best1$sig_thresh[1] else 0.010
best_mag <- if (nrow(best1) > 0) best1$mag_thresh[1] else 0.080

grid2 <- tidyr::expand_grid(
  lookback_roll = seq(40, 85, by = 5),
  sma_window    = c(5, 8, 10, 13, 15, 18, 20, 23, 25, 30)
)

cl2 <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl2)

parallel::clusterExport(cl2, varlist = c(
  "run_strategy",
  "sector_train", "spy_train", "prices_wide", "div_wide",
  "cluster_daily", "sector_tickers",
  "max_dd", "apply_kf", "get_alloc", "omega_ratio", "calmar", "ann_ret",
  "TRAIN_END", "STRESS_DAYS", "KF_DW_SCALE", "INITIAL_CAPITAL", "TC_RATE",
  "best_sig", "best_mag"
), envir = globalenv())

invisible(parallel::clusterEvalQ(cl2, {
  library(tidyverse); library(dlm); library(TTR); library(slider); library(lubridate)
}))

opt2 <- foreach::foreach(
  i         = seq_len(nrow(grid2)),
  .combine  = dplyr::bind_rows,
  .packages = c("dplyr", "tidyr", "purrr", "lubridate", "TTR", "dlm", "slider")
) %dopar% {
  run_strategy(
    lookback_roll = grid2$lookback_roll[i],
    sma_window    = grid2$sma_window[i],
    sig_thresh    = best_sig,
    mag_thresh    = best_mag
  )
}

parallel::stopCluster(cl2)
```

### 9.3 Optimization Surfaces

```{r opt-surface1}
#| message: false
#| warning: false
#| echo: false

par_sig <- sort(unique(opt1$sig_thresh))
par_mag <- sort(unique(opt1$mag_thresh))

z_omega1 <- opt1 %>%
  dplyr::select(sig_thresh, mag_thresh, omega) %>%
  tidyr::pivot_wider(names_from = mag_thresh, values_from = omega) %>%
  dplyr::select(-sig_thresh) %>%
  as.matrix()

plotly::plot_ly(
  x = ~par_mag,
  y = ~par_sig,
  z = ~z_omega1
) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(
      text = "Grid 1: Omega Ratio by Signal and Magnitude Thresholds<br><sup>Lookback = 60 days, SMA = 20 days. Training period only.</sup>",
      x = 0.5
    ),
    scene = list(
      xaxis = list(title = "Magnitude Threshold"),
      yaxis = list(title = "Signal Threshold"),
      zaxis = list(title = "Omega Ratio")
    )
  )
```

Grid 1 shows a broad elevated ridge at signal threshold 0.010 to 0.050 and magnitude threshold 0.050 to 0.080, with Omega staying above 1.12 across the entire region. The peak at sig = 0.050 and mag = 0.070 is not a sharp isolated spike; the surrounding cells are nearly as strong, which means the strategy is not dependent on hitting an exact parameter value. This plateau shape is the target outcome: it indicates the edge is structural rather than coincidental. Performance collapses at high signal thresholds (0.060 to 0.080) because the strategy almost never fires a momentum signal, and at very low magnitude thresholds (0.010 to 0.030) because the stress state is too easily triggered, parking the portfolio in cash during periods that did not warrant it.

```{r opt-surface2}
#| message: false
#| warning: false
#| echo: false

par_lookback <- sort(unique(opt2$lookback_roll))
par_sma      <- sort(unique(opt2$sma_window))

z_omega2 <- opt2 %>%
  dplyr::select(lookback_roll, sma_window, omega) %>%
  tidyr::pivot_wider(names_from = sma_window, values_from = omega) %>%
  dplyr::select(-lookback_roll) %>%
  as.matrix()

plotly::plot_ly(
  x = ~par_sma,
  y = ~par_lookback,
  z = ~z_omega2
) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(
      text = paste0("Grid 2: Omega Ratio by Lookback and SMA Window<br>",
                    "<sup>Signal threshold = ", best_sig,
                    ", Magnitude threshold = ", best_mag,
                    ". Training period only.</sup>"),
      x = 0.5
    ),
    scene = list(
      xaxis = list(title = "SMA Window (days)"),
      yaxis = list(title = "Lookback Window (days)"),
      zaxis = list(title = "Omega Ratio")
    )
  )
```

Grid 2 is noisier than Grid 1, which confirms that window parameters drive less of the performance difference than thresholds do. The selected lookback of 50 days and SMA of 23 days sit in an elevated region but the surface has more variance around them. The 3-day difference between SMA = 20 and SMA = 23 is not meaningful in isolation; the takeaway is that the strategy is not sensitive to small changes in window length. The slightly shorter lookback of 50 days versus the default of 60 makes intuitive sense: it allows the clustering features to respond to recent regime changes without sacrificing stability.

### 9.4 Multi-Metric Z-Score Heatmaps

A single-metric optimization surface (Omega alone) risks selecting parameters that win on one dimension while quietly losing on others. The Z-score heatmap normalizes every metric to a common scale -- $(x - \mu) / \sigma$ across the grid -- and plots all of them simultaneously. Blue cells are above-average for that metric, red cells are below-average. A parameter combination that is consistently blue across all panels is a genuinely robust choice. A combination that peaks on Omega but turns red on drawdown or hit rate is a warning sign. For max drawdown, a higher (less negative) value is better, so blue still means good.

```{r opt-zscore1}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 11
#| fig-height: 9

metrics_long1 <- opt1 %>%
  dplyr::select(sig_thresh, mag_thresh,
                Omega = omega, Calmar = calmar_r, Sharpe = sharpe,
                Sortino = sortino, `Ann. Return` = ann_ret_pct,
                `Max DD` = max_dd_pct, `Hit Rate` = hit_rate) %>%
  tidyr::pivot_longer(
    cols      = -c(sig_thresh, mag_thresh),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  dplyr::group_by(metric) %>%
  dplyr::mutate(valueZ = (value - mean(value, na.rm = TRUE)) /
                          sd(value, na.rm = TRUE)) %>%
  dplyr::ungroup()

ggplot(metrics_long1, aes(x = sig_thresh, y = mag_thresh, fill = valueZ)) +
  geom_raster(interpolate = TRUE) +
  facet_wrap(~metric, scales = "free") +
  scale_fill_gradient2(
    low      = "red",
    mid      = "white",
    high     = "blue",
    midpoint = 0,
    name     = "Z-Score"
  ) +
  scale_x_continuous(labels = scales::number_format(accuracy = 0.01)) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.01)) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid   = element_blank(),
    strip.text   = element_text(face = "bold")
  ) +
  labs(
    title    = "Grid 1: Multi-Metric Z-Score Heatmap (Thresholds)",
    subtitle = "Blue = above average for that metric. Consistent blue across panels = robust parameter combination.",
    x        = "Signal Threshold",
    y        = "Magnitude Threshold",
    caption  = "Lookback = 60 days, SMA = 20 days. Training period only."
  )
```

```{r opt-zscore2}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 11
#| fig-height: 9

metrics_long2 <- opt2 %>%
  dplyr::select(lookback_roll, sma_window,
                Omega = omega, Calmar = calmar_r, Sharpe = sharpe,
                Sortino = sortino, `Ann. Return` = ann_ret_pct,
                `Max DD` = max_dd_pct, `Hit Rate` = hit_rate) %>%
  tidyr::pivot_longer(
    cols      = -c(lookback_roll, sma_window),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  dplyr::group_by(metric) %>%
  dplyr::mutate(valueZ = (value - mean(value, na.rm = TRUE)) /
                          sd(value, na.rm = TRUE)) %>%
  dplyr::ungroup()

ggplot(metrics_long2, aes(x = sma_window, y = lookback_roll, fill = valueZ)) +
  geom_raster(interpolate = TRUE) +
  facet_wrap(~metric, scales = "free") +
  scale_fill_gradient2(
    low      = "red",
    mid      = "white",
    high     = "blue",
    midpoint = 0,
    name     = "Z-Score"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  ) +
  labs(
    title    = "Grid 2: Multi-Metric Z-Score Heatmap (Windows)",
    subtitle = "Blue = above average for that metric. Consistent blue across panels = robust parameter combination.",
    x        = "SMA Window (days)",
    y        = "Lookback Window (days)",
    caption  = paste0("Signal threshold = ", best_sig, ", Magnitude threshold = ", best_mag, ". Training period only.")
  )
```

### 9.5 Selected Parameters

The Z-score heatmaps provide stronger evidence for the selected parameters than the 3D surfaces alone. In Grid 1, a broad blue band spanning signal threshold 0.020 to 0.050 appears consistently across Ann. Return, Calmar, Omega, Sharpe, and Sortino simultaneously. The selected sig=0.050 sits at the right edge of that band and mag=0.070 sits in the middle of the blue region across the magnitude axis. Five independent metrics pointing to the same parameter region confirms the edge is structural rather than an artifact of optimizing a single statistic. The one exception is Hit Rate, which favors very low signal thresholds where the strategy is invested nearly every day. This is expected and not a concern: high hit rate at low thresholds reflects a flatter, more active strategy, not a better one, and the tail-risk metrics correctly penalize that region. In Grid 2, the SMA window shows the clearest and most consistent signal: a blue band at SMA 15 to 25 runs across almost every metric panel, placing the selected SMA=23 firmly in a robust region. The lookback dimension is noisier and less decisive, which is the honest read: lookback at 50 days is a marginal improvement over the default 60, but the difference is small enough that the default would not have been a meaningful mistake.

The final parameters are selected from the intersection of a high-Omega region and a Calmar Ratio above 0.10, which filters out combinations that inflate returns at the cost of large drawdowns. Parameters near the edge of the search range are treated with skepticism: a peak at the boundary suggests the optimum lies outside the grid and the apparent result may not generalize.

```{r opt-selected}
#| message: false
#| warning: false
#| echo: false

# Grid 1 best: thresholds (max Omega, Calmar > 0.10)
best_g1 <- opt1 %>%
  dplyr::filter(calmar_r > 0.10) %>%
  dplyr::slice_max(omega, n = 1)

# Grid 2 best: windows (max Omega, Calmar > 0.10)
best_g2 <- opt2 %>%
  dplyr::filter(calmar_r > 0.10) %>%
  dplyr::slice_max(omega, n = 1)

final_params <- tibble::tibble(
  Parameter   = c("Signal Threshold", "Magnitude Threshold",
                  "Lookback Window",  "SMA Window"),
  Default     = c(0.010, 0.080, 60, 20),
  Selected    = c(
    if (nrow(best_g1) > 0) best_g1$sig_thresh[1]    else 0.010,
    if (nrow(best_g1) > 0) best_g1$mag_thresh[1]    else 0.080,
    if (nrow(best_g2) > 0) best_g2$lookback_roll[1] else 60,
    if (nrow(best_g2) > 0) best_g2$sma_window[1]    else 20
  ),
  `Omega at Selected` = c(
    if (nrow(best_g1) > 0) round(best_g1$omega[1], 3) else NA,
    if (nrow(best_g1) > 0) round(best_g1$omega[1], 3) else NA,
    if (nrow(best_g2) > 0) round(best_g2$omega[1], 3) else NA,
    if (nrow(best_g2) > 0) round(best_g2$omega[1], 3) else NA
  ),
  `Calmar at Selected` = c(
    if (nrow(best_g1) > 0) round(best_g1$calmar_r[1], 3) else NA,
    if (nrow(best_g1) > 0) round(best_g1$calmar_r[1], 3) else NA,
    if (nrow(best_g2) > 0) round(best_g2$calmar_r[1], 3) else NA,
    if (nrow(best_g2) > 0) round(best_g2$calmar_r[1], 3) else NA
  )
)

final_params %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Optimization Results: Selected Parameters",
    subtitle = "Objective: maximize Omega Ratio. Filter: Calmar Ratio > 0.10. Training period only."
  ) %>%
  gt::fmt_number(columns = c(Default, Selected), decimals = 3) %>%
  gt::fmt_number(columns = c(`Omega at Selected`, `Calmar at Selected`), decimals = 3) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#e8f5e9"),
    locations = gt::cells_body(columns = Selected)
  ) %>%
  gt::tab_source_note("Parameters fit on training data only. No refitting on testing period.")
```

The optimization is intentionally constrained. With only 11 ETFs and two clusters, the strategy has limited degrees of freedom, so over-fitting is a real risk on a 24-year training window. The Calmar filter is designed to penalize combinations that happen to catch a strong run while concealing a destructive drawdown. The goal is a parameter set that sits in a stable plateau on the surface, not a single sharp peak. A combination at the centre of a broad high-Omega region generalizes better than one at a narrow spike, even if the spike appears higher on the training surface.

### 9.6 COVID Robustness Check

A legitimate concern with any optimization over the 2000-2024 training period is that the winning parameters may simply be those that happened to avoid the COVID crash of February to April 2020. That event produced the sharpest drawdown in the dataset and would disproportionately reward any parameter set that kept the strategy in cash during those months. To test whether the selected parameters are genuinely better or just COVID-lucky, the threshold grid is re-run with the COVID stress period excluded from performance measurement. If the same parameters win on both surfaces, the selection is robust. If different parameters win, the original result is partly an artifact of a single event.

```{r opt-covid-robust}
#| message: false
#| warning: false
#| echo: false
#| cache: true

COVID_EXCL_START <- "2020-02-15"
COVID_EXCL_END   <- "2020-06-30"

cl3 <- parallel::makeCluster(n_cores)
doParallel::registerDoParallel(cl3)

parallel::clusterExport(cl3, varlist = c(
  "run_strategy",
  "sector_train", "spy_train", "prices_wide", "div_wide",
  "cluster_daily", "sector_tickers",
  "max_dd", "apply_kf", "get_alloc", "omega_ratio", "calmar", "ann_ret",
  "TRAIN_END", "STRESS_DAYS", "KF_DW_SCALE", "INITIAL_CAPITAL", "TC_RATE",
  "COVID_EXCL_START", "COVID_EXCL_END"
), envir = globalenv())

invisible(parallel::clusterEvalQ(cl3, {
  library(tidyverse); library(dlm); library(TTR); library(slider); library(lubridate)
}))

opt_covid <- foreach::foreach(
  i         = seq_len(nrow(grid1)),
  .combine  = dplyr::bind_rows,
  .packages = c("dplyr", "tidyr", "purrr", "lubridate", "TTR", "dlm", "slider")
) %dopar% {
  run_strategy(
    lookback_roll = 60,
    sma_window    = 20,
    sig_thresh    = grid1$sig_thresh[i],
    mag_thresh    = grid1$mag_thresh[i],
    excl_start    = COVID_EXCL_START,
    excl_end      = COVID_EXCL_END
  )
}

parallel::stopCluster(cl3)
```

```{r opt-covid-surface}
#| message: false
#| warning: false
#| echo: false

par_sig_c <- sort(unique(opt_covid$sig_thresh))
par_mag_c <- sort(unique(opt_covid$mag_thresh))

z_covid <- opt_covid %>%
  dplyr::select(sig_thresh, mag_thresh, omega) %>%
  tidyr::pivot_wider(names_from = mag_thresh, values_from = omega) %>%
  dplyr::select(-sig_thresh) %>%
  as.matrix()

plotly::plot_ly(
  x = ~par_mag_c,
  y = ~par_sig_c,
  z = ~z_covid
) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(
      text = "COVID Robustness: Omega Ratio Excluding Feb-Jun 2020<br><sup>Same threshold grid as Grid 1. Lookback = 60 days, SMA = 20 days.</sup>",
      x = 0.5
    ),
    scene = list(
      xaxis = list(title = "Magnitude Threshold"),
      yaxis = list(title = "Signal Threshold"),
      zaxis = list(title = "Omega Ratio")
    )
  )
```

```{r opt-covid-comparison}
#| message: false
#| warning: false
#| echo: false

best_covid <- opt_covid %>%
  dplyr::filter(calmar_r > 0.10) %>%
  dplyr::slice_max(omega, n = 1)

dplyr::bind_rows(
  opt1 %>%
    dplyr::filter(calmar_r > 0.10) %>%
    dplyr::slice_max(omega, n = 1) %>%
    dplyr::mutate(Dataset = "Full Training (incl. COVID)"),
  best_covid %>%
    dplyr::mutate(Dataset = "COVID Excluded (Feb-Jun 2020)")
) %>%
  dplyr::select(Dataset, sig_thresh, mag_thresh, omega, calmar_r) %>%
  dplyr::rename(
    `Signal Threshold`    = sig_thresh,
    `Magnitude Threshold` = mag_thresh,
    `Omega Ratio`         = omega,
    `Calmar Ratio`        = calmar_r
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "COVID Robustness: Optimal Thresholds With vs. Without COVID",
    subtitle = "If selected parameters are consistent, the optimization is not COVID-driven."
  ) %>%
  gt::fmt_number(columns = c(`Signal Threshold`, `Magnitude Threshold`), decimals = 3) %>%
  gt::fmt_number(columns = c(`Omega Ratio`, `Calmar Ratio`), decimals = 3) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_source_note("Window parameters fixed at defaults for comparability. COVID period defined as Feb 15 to Jun 30, 2020.")
```

The COVID robustness result is the strongest possible outcome: optimal parameters are identical to the decimal place, and Omega changes by 0.001. Removing the sharpest drawdown event in the dataset makes essentially no difference to which parameter combination wins. This directly addresses a failure from the FIN 440 quantitative trading project, where the strategy was unknowingly optimized against the COVID period. The FIN 440 model identified a set of parameters that happened to keep the portfolio in cash during February to April 2020, which produced excellent training metrics but generated almost no return in the testing period because the strategy had no genuine edge -- **it was a COVID avoidance model, not a trading model.** The design choices here explicitly guard against that outcome. The stress state is triggered by a persistent signal requiring three consecutive days of confirmation, not a single spike. The Omega objective rewards consistent gains over losses across all market conditions, not just survival of the worst month. And the COVID exclusion test confirms the edge exists in the other 23 years of data, not only in the four months the FIN 440 model was implicitly built around.

------------------------------------------------------------------------

## 10. Testing Period Results

The testing period runs from January 2025 to March 2026, approximately 15 months. All parameters are fixed at the values selected from the training optimization. No refitting occurs. The rolling cluster assignments are extended into the test period using the same beta-anchored algorithm applied during training, with each month-end cluster computed using only data available up to that point. The Kalman Filter and SMA are computed on the full cumulative return series running from training start through the test period, maintaining the causal forward-pass property. The strategy is re-initialized at \$1,000,000 at the start of the test period as a clean independent evaluation.

```{r test-prep}
#| message: false
#| warning: false
#| echo: false
#| cache: true

# Optimized parameters from Section 9
OPT_SIG      <- if (nrow(best_g1) > 0) best_g1$sig_thresh[1]    else 0.010
OPT_MAG      <- if (nrow(best_g1) > 0) best_g1$mag_thresh[1]    else 0.080
OPT_LOOKBACK <- if (nrow(best_g2) > 0) best_g2$lookback_roll[1] else 60
OPT_SMA      <- if (nrow(best_g2) > 0) best_g2$sma_window[1]    else 20

sector_full <- returns %>% dplyr::filter(symbol %in% sector_tickers)
spy_full    <- returns %>% dplyr::filter(symbol == "SPY")

# ── Extend cluster_daily to full period using static assignments ─────────────
cluster_daily_full <- returns %>%
  dplyr::filter(symbol %in% sector_tickers) %>%
  dplyr::left_join(
    cluster_assignments_spy %>% dplyr::select(symbol, cluster),
    by = "symbol"
  ) %>%
  dplyr::group_by(date, cluster) %>%
  dplyr::summarise(cluster_ret = mean(ret, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(cluster_label = dplyr::case_when(
    cluster == cyclical_id  ~ "Cyclical",
    cluster == defensive_id ~ "Defensive"
  ))

# ── KF + SMA on full cumulative return series ────────────────────────────────
cluster_signals_full <- cluster_daily_full %>%
  dplyr::group_by(cluster_label) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    cum_ret  = cumsum(cluster_ret),
    kf_level = apply_kf(cum_ret, dw_scale = KF_DW_SCALE),
    sma_n    = TTR::SMA(cum_ret, n = OPT_SMA)
  ) %>%
  dplyr::ungroup()

# ── Signals for full period with optimized thresholds ────────────────────────
signal_full <- cluster_signals_full %>%
  dplyr::filter(!is.na(sma_n)) %>%
  dplyr::mutate(spread = sma_n - kf_level) %>%
  dplyr::select(date, cluster_label, spread) %>%
  tidyr::pivot_wider(names_from = cluster_label, values_from = spread) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    stress_raw = (Cyclical < 0 & Defensive < 0) |
                 abs(Cyclical) > OPT_MAG        |
                 abs(Defensive) > OPT_MAG,
    stress_confirmed = slider::slide_lgl(
      stress_raw, ~all(.x), .before = STRESS_DAYS - 1, .complete = TRUE
    ),
    stress_confirmed = tidyr::replace_na(stress_confirmed, FALSE),
    signal = dplyr::case_when(
      stress_confirmed                                  ~ "Stress",
      Cyclical  > OPT_SIG & Cyclical  >= Defensive     ~ "Momentum Cyclical",
      Defensive > OPT_SIG & Defensive >  Cyclical      ~ "Momentum Defensive",
      TRUE                                             ~ "No Signal"
    )
  )

# ── Extend rolling clusters into test period ─────────────────────────────────
month_ends_test <- sector_full %>%
  dplyr::filter(symbol == "XLK", date >= TEST_START) %>%
  dplyr::mutate(ym = lubridate::floor_date(date, "month")) %>%
  dplyr::group_by(ym) %>%
  dplyr::slice_max(date, n = 1) %>%
  dplyr::ungroup() %>%
  dplyr::pull(date)

cluster_window_full <- function(end_date) {
  ws   <- sector_full %>%
    dplyr::filter(date <= end_date) %>%
    dplyr::group_by(symbol) %>%
    dplyr::slice_tail(n = OPT_LOOKBACK) %>%
    dplyr::ungroup()
  wspy <- spy_full %>%
    dplyr::filter(date <= end_date) %>%
    dplyr::slice_tail(n = OPT_LOOKBACK)

  sc  <- sum(wspy$ret, na.rm = TRUE)
  sv  <- var(wspy$ret, na.rm = TRUE)
  smd <- max_dd(wspy$ret)

  feats <- ws %>%
    dplyr::group_by(symbol) %>%
    dplyr::summarise(
      excess_ret = sum(ret, na.rm = TRUE) - sc,
      beta       = cov(ret, wspy$ret[seq_len(dplyr::n())],
                       use = "complete.obs") / sv,
      mdd_ratio  = max_dd(ret) / smd,
      .groups    = "drop"
    )

  mat  <- scale(feats %>% dplyr::select(excess_ret, beta, mdd_ratio))
  rownames(mat) <- feats$symbol
  labs <- cutree(hclust(dist(mat, method = "euclidean"), method = "ward.D2"), k = 2)
  mb   <- tapply(feats$beta, labs, mean)
  cr   <- as.integer(names(which.max(mb)))

  tibble::tibble(
    date    = end_date,
    symbol  = feats$symbol,
    cluster = ifelse(labs == cr, "Cyclical", "Defensive")
  )
}

rolling_test     <- purrr::map_dfr(month_ends_test, cluster_window_full)
rolling_combined <- dplyr::bind_rows(rolling_spy, rolling_test)

all_dates_full <- returns %>%
  dplyr::filter(symbol == "XLK") %>%
  dplyr::arrange(date) %>%
  dplyr::select(date)

rolling_filled_full <- tidyr::expand_grid(
  date   = all_dates_full$date,
  symbol = sector_tickers
) %>%
  dplyr::left_join(
    rolling_combined %>% dplyr::select(date, symbol, cluster),
    by = c("date", "symbol")
  ) %>%
  dplyr::arrange(symbol, date) %>%
  dplyr::group_by(symbol) %>%
  tidyr::fill(cluster, .direction = "down") %>%
  dplyr::ungroup() %>%
  dplyr::mutate(cluster = dplyr::if_else(cluster == "Cyclical", 1L, 0L))

cluster_wide_test <- rolling_filled_full %>%
  tidyr::pivot_wider(names_from  = symbol,
                     values_from = cluster,
                     names_prefix = "clust_")

div_wide_full <- dividends_raw %>%
  tidyr::pivot_wider(
    names_from   = symbol,
    values_from  = dividend,
    names_prefix = "div_",
    values_fill  = 0
  )

prices_wide_full <- raw %>%
  dplyr::filter(symbol %in% sector_tickers) %>%
  dplyr::select(symbol, date, open, close) %>%
  tidyr::pivot_wider(
    names_from  = symbol,
    values_from = c(open, close),
    names_glue  = "{symbol}_{.value}"
  ) %>%
  dplyr::arrange(date)

loop_test <- prices_wide_full %>%
  dplyr::left_join(div_wide_full,   by = "date") %>%
  dplyr::left_join(cluster_wide_test, by = "date") %>%
  dplyr::left_join(signal_full %>% dplyr::select(date, signal), by = "date") %>%
  dplyr::arrange(date) %>%
  dplyr::filter(!is.na(signal), date >= TEST_START) %>%
  dplyr::mutate(dplyr::across(dplyr::starts_with("div_"),
                               ~tidyr::replace_na(.x, 0))) %>%
  dplyr::mutate(ym = lubridate::floor_date(date, "month")) %>%
  dplyr::group_by(ym) %>%
  dplyr::mutate(month_end_flag = date == max(date)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-ym)
```

```{r test-execution-loop}
#| message: false
#| warning: false
#| echo: false
#| cache: true

n_test     <- nrow(loop_test)
pv_test_h  <- numeric(n_test)
cash_test_h <- numeric(n_test)
tc_test_h  <- numeric(n_test)
sig_test_h <- character(n_test)
w_test_h   <- matrix(0, nrow = n_test, ncol = length(sector_tickers),
                      dimnames = list(NULL, sector_tickers))

# Seed initial signal from last active state in training period
last_train_sig <- signal_full %>%
  dplyr::filter(date < as.Date(TEST_START), signal != "No Signal") %>%
  dplyr::slice_tail(n = 1) %>%
  dplyr::pull(signal)

cur_cash       <- INITIAL_CAPITAL
cur_shares     <- setNames(rep(0, length(sector_tickers)), sector_tickers)
pend_sig       <- if (length(last_train_sig) > 0) last_train_sig else NA_character_
cur_active_sig <- pend_sig   # tracks last executed signal for month-end rebalancing
prev_pv        <- INITIAL_CAPITAL

for (i in seq_len(n_test)) {
  td  <- loop_test[i, ]
  tds <- as.character(td$signal)

  div_inc <- sum(sapply(sector_tickers, function(etf) {
    dc <- paste0("div_", etf)
    if (dc %in% names(td)) { dv <- td[[dc]]; if (!is.na(dv) && dv > 0) return(cur_shares[[etf]] * dv) }
    0
  }), na.rm = TRUE)
  cur_cash <- cur_cash + div_inc

  tc <- 0
  # Rebalance on signal change OR month-end (captures shifting cluster membership)
  is_me_t <- isTRUE(loop_test$month_end_flag[i])
  exec_sig_t <- if (!is.na(pend_sig) && pend_sig != "No Signal") {
    pend_sig
  } else if (is_me_t && !is.na(cur_active_sig) && cur_active_sig != "No Signal") {
    cur_active_sig
  } else {
    NA_character_
  }

  if (!is.na(exec_sig_t)) {
    alloc <- get_alloc(exec_sig_t)
    ct <- setNames(sapply(sector_tickers, function(etf) {
      col <- paste0("clust_", etf); if (col %in% names(td)) td[[col]] else NA_integer_
    }), sector_tickers)
    ac <- sector_tickers[!is.na(ct) & ct == 1 &
            !is.na(sapply(sector_tickers, function(e) td[[paste0(e, "_open")]]))]
    ad <- sector_tickers[!is.na(ct) & ct == 0 &
            !is.na(sapply(sector_tickers, function(e) td[[paste0(e, "_open")]]))]
    tw <- setNames(rep(0, length(sector_tickers)), sector_tickers)
    if (length(ac) > 0) tw[ac] <- alloc$cyc / length(ac)
    if (length(ad) > 0) tw[ad] <- alloc$def / length(ad)
    tdols <- prev_pv * tw
    ns    <- cur_shares
    for (etf in sector_tickers) {
      op <- td[[paste0(etf, "_open")]]; if (is.na(op) || op <= 0) next
      ns[[etf]]  <- tdols[[etf]] / op
      dlt        <- ns[[etf]] - cur_shares[[etf]]
      cur_cash   <- cur_cash - dlt * op
      tc         <- tc + abs(dlt * op) * TC_RATE
    }
    cur_cash   <- cur_cash - tc
    cur_shares <- ns
    if (!is.na(pend_sig) && pend_sig != "No Signal")
      cur_active_sig <- pend_sig
  }

  eq <- sum(sapply(sector_tickers, function(etf) {
    cp <- td[[paste0(etf, "_close")]]; if (!is.na(cp) && cp > 0) cur_shares[[etf]] * cp else 0
  }), na.rm = TRUE)

  pv <- eq + cur_cash
  pv_test_h[i]  <- pv
  cash_test_h[i] <- cur_cash
  tc_test_h[i]  <- tc
  sig_test_h[i] <- tds

  if (pv > 0) {
    for (etf in sector_tickers) {
      cp <- td[[paste0(etf, "_close")]]
      if (!is.na(cp) && cp > 0)
        w_test_h[i, etf] <- (cur_shares[[etf]] * cp) / pv
    }
  }

  ps       <- if (i == 1) NA_character_ else sig_test_h[i - 1]
  pend_sig <- dplyr::case_when(
    is.na(ps)  & tds != "No Signal"             ~ tds,
    !is.na(ps) & tds != ps & tds != "No Signal" ~ tds,
    TRUE                                         ~ NA_character_
  )
  prev_pv <- pv
}

test_results <- loop_test %>%
  dplyr::select(date) %>%
  dplyr::mutate(
    portfolio_value = pv_test_h,
    cash            = cash_test_h,
    tc              = tc_test_h,
    signal          = sig_test_h
  )

test_strat_r <- diff(test_results$portfolio_value) / head(test_results$portfolio_value, -1)
test_strat_r <- test_strat_r[is.finite(test_strat_r)]
```

### 10.1 Strategy Results Chart

```{r test-xts-chart}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| fig-keep: last

tick_test <- test_results %>%
  dplyr::mutate(
    strat_idx  = portfolio_value / dplyr::first(portfolio_value) * 100,
    trade_flag = dplyr::if_else(tc > 0, 1L, 0L),
    pos_num    = dplyr::case_when(
      signal == "Momentum Cyclical"  ~  1,
      signal == "Momentum Defensive" ~ -1,
      signal == "Stress"             ~ -2,
      TRUE                           ~  0
    )
  ) %>%
  dplyr::select(date, strat_idx, trade_flag, pos_num) %>%
  timetk::tk_xts(date_var = date)

plot(tick_test$strat_idx,
     main          = "Strategy Results",
     ylab          = "Indexed to 100",
     col           = "black",
     lwd           = 1.2,
     grid.ticks.on = "months")

xts::addSeries(
  tick_test$trade_flag,
  main = "Trades",
  on   = NA,
  type = "h",
  col  = "#2c7bb6",
  lwd  = 1,
  pch  = 0
)

xts::addSeries(
  tick_test$pos_num,
  main = "Position (1=Cyc, -1=Def, -2=Stress)",
  on   = NA,
  type = "h",
  col  = "#2c7bb6",
  lwd  = 1,
  pch  = 0
)

xts::addSeries(
  tick_test$strat_idx,
  main = "CumEQ",
  on   = NA,
  type = "l",
  col  = "#2c7bb6",
  lwd  = 1,
  pch  = 0
)
```

The four-panel chart reveals the most important limitation of this strategy in the testing period: it executed only one trade in 15 months. The Trades panel shows a single spike around late April 2025, coinciding with the tariff-driven market selloff, at which point the strategy rotated into a defensive or stress position as visible in the Position panel. After that episode resolved, the strategy returned to No Signal and held that position for the remainder of the test window without generating another rebalancing event. This is a direct consequence of the signal threshold being set at 0.050: while this level was selected by maximizing Omega Ratio on the training surface, the realized spread between the Cyclical and Defensive clusters rarely crossed that threshold in the relatively low-volatility environment of mid-to-late 2025. A strategy that trades once in 15 months is not functioning as a dynamic rotation engine; it is functioning as a one-decision portfolio that happened to make the right call during one stress event. The risk-adjusted metrics look favorable in part because the strategy avoided the April 2025 drawdown, but the honest interpretation is that a single correct signal does not validate the full architecture. It is insufficient evidence that the clustering and Kalman filter machinery is producing repeatable, actionable signals across a range of market conditions.

### 10.2 Strategy vs SPY Equity Curve

```{r test-equity-curve}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 5

spy_test_bench <- raw %>%
  dplyr::filter(symbol == "SPY", date >= TEST_START) %>%
  dplyr::select(date, spy_adj = adjusted) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    spy_ret   = spy_adj / dplyr::lag(spy_adj) - 1,
    spy_value = INITIAL_CAPITAL * (spy_adj / dplyr::first(spy_adj))
  )

test_spy_r <- spy_test_bench$spy_ret[!is.na(spy_test_bench$spy_ret)]

# Equal-weight sector ETF benchmark: compute returns within each symbol, then average by date
ew_bench <- raw %>%
  dplyr::filter(symbol %in% sector_tickers, date >= TEST_START) %>%
  dplyr::group_by(symbol) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(ret = adjusted / dplyr::lag(adjusted) - 1) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(ew_ret = mean(ret, na.rm = TRUE), .groups = "drop") %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    ew_ret   = dplyr::if_else(is.nan(ew_ret), 0, ew_ret),
    ew_value = INITIAL_CAPITAL * cumprod(1 + ew_ret)
  )

test_ew_r <- ew_bench$ew_ret

# XLK benchmark
xlk_bench <- raw %>%
  dplyr::filter(symbol == "XLK", date >= TEST_START) %>%
  dplyr::select(date, xlk_adj = adjusted) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(
    xlk_ret   = xlk_adj / dplyr::lag(xlk_adj) - 1,
    xlk_value = INITIAL_CAPITAL * (xlk_adj / dplyr::first(xlk_adj))
  )

test_xlk_r <- xlk_bench$xlk_ret[!is.na(xlk_bench$xlk_ret)]

combined_test <- test_results %>%
  dplyr::left_join(spy_test_bench %>% dplyr::select(date, spy_value),  by = "date") %>%
  dplyr::left_join(ew_bench       %>% dplyr::select(date, ew_value),   by = "date") %>%
  dplyr::left_join(xlk_bench      %>% dplyr::select(date, xlk_value),  by = "date") %>%
  dplyr::mutate(
    strat_idx  = portfolio_value / dplyr::first(portfolio_value) * 100,
    spy_idx    = spy_value  / dplyr::first(spy_value)  * 100,
    ew_idx     = ew_value   / dplyr::first(ew_value)   * 100,
    xlk_idx    = xlk_value  / dplyr::first(xlk_value)  * 100,
    strat_peak = cummax(portfolio_value),
    spy_peak   = cummax(spy_value),
    ew_peak    = cummax(ew_value),
    strat_dd   = (portfolio_value - strat_peak) / strat_peak,
    spy_dd     = (spy_value - spy_peak) / spy_peak,
    ew_dd      = (ew_value - ew_peak) / ew_peak
  )

combined_test %>%
  dplyr::select(date,
                Strategy       = strat_idx,
                SPY            = spy_idx,
                `Equal-Weight` = ew_idx,
                XLK            = xlk_idx) %>%
  tidyr::pivot_longer(-date, names_to = "series", values_to = "value") %>%
  ggplot(aes(x = date, y = value, colour = series, linewidth = series)) +
  geom_line(alpha = 0.9) +
  scale_colour_manual(values = c(
    "Strategy"     = "#2c7bb6",
    "SPY"          = "#d7191c",
    "Equal-Weight" = "#1a9641",
    "XLK"          = "#f4a736"
  )) +
  scale_linewidth_manual(values = c(
    "Strategy"     = 1.0,
    "SPY"          = 0.6,
    "Equal-Weight" = 0.6,
    "XLK"          = 0.6
  )) +
  labs(
    title    = "Testing Period: Strategy vs Benchmarks (Indexed to 100)",
    subtitle = "January 2025 to March 2026. Optimized parameters applied. No refitting.",
    x        = NULL, y = "Index (100 = start)", colour = NULL, linewidth = NULL,
    caption  = "Equal-Weight = buy-and-hold equal allocation across all 11 sector ETFs. XLK = Technology ETF."
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The strategy underperformed all three benchmarks on total return over the 15-month test window, finishing at 6.5% versus SPY at 9.7%, Equal-Weight at 14.7%, and XLK at 10.8%. The honest explanation for most of that gap is that the technology sector had an exceptional run driven by AI investment themes, and SPY, Equal-Weight, and XLK all carried meaningful tech exposure. This strategy did not overweight technology, which is the correct decision. Adding XLK exposure in hindsight would be speculative, not systematic, and would defeat the purpose of a rule-based signal. The strategy was not designed to chase sector momentum; it was designed to rotate between two structurally distinct clusters with risk control built in. Judged on that mandate, the equity curve tells a story of lower participation during the bull run but also visibly less damage during the April 2025 tariff shock and more stability through the volatile late-2025 period when all three benchmarks gave back gains.

### 10.3 Drawdown Comparison

```{r test-drawdown}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 4

combined_test %>%
  dplyr::select(date, Strategy = strat_dd, SPY = spy_dd, `Equal-Weight` = ew_dd) %>%
  tidyr::pivot_longer(-date, names_to = "series", values_to = "dd") %>%
  ggplot(aes(x = date, y = dd * 100, colour = series, fill = series)) +
  geom_ribbon(aes(ymin = dd * 100, ymax = 0), alpha = 0.15) +
  geom_line(linewidth = 0.6) +
  scale_colour_manual(values = c(
    "Strategy"     = "#2c7bb6",
    "SPY"          = "#d7191c",
    "Equal-Weight" = "#1a9641"
  )) +
  scale_fill_manual(values = c(
    "Strategy"     = "#2c7bb6",
    "SPY"          = "#d7191c",
    "Equal-Weight" = "#1a9641"
  )) +
  labs(
    title    = "Testing Period: Drawdown Comparison",
    subtitle = "Drawdown from rolling peak. January 2025 to March 2026.",
    x        = NULL, y = "Drawdown (%)", colour = NULL, fill = NULL,
    caption  = "Testing period only. Parameters fixed from training optimization."
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

The drawdown chart is the clearest evidence in favour of the strategy's design. During the April 2025 tariff shock, SPY fell approximately 18% from peak and Equal-Weight fell roughly 15%, while the strategy's maximum drawdown was -5.6% across the entire test window. The stress detection mechanism appears to have done meaningful work during that episode, reducing equity exposure before or during the sharpest phase of the selloff. The strategy also held its value through the late-2025 period where the other benchmarks were flat to negative, which is visible as the blue line staying elevated while red and green declined. These are the conditions the stress state was designed for, and the drawdown chart validates that the mechanism is functioning as intended.

### 10.4 Performance Metrics

```{r test-performance-table}
#| message: false
#| warning: false
#| echo: false

bench_metrics <- function(r, start_val, end_val) {
  c(
    round((end_val / start_val - 1) * 100, 2),
    round(ann_ret(r) * 100, 2),
    round(mean(r, na.rm = TRUE) / sd(r, na.rm = TRUE) * sqrt(252), 3),
    round(omega_ratio(r), 3),
    round(mean(r[r > 0], na.rm = TRUE) /
            abs(mean(r[r < 0], na.rm = TRUE)) * sqrt(252), 3),
    round(max_dd(r) * 100, 2),
    round(calmar(r), 3),
    round(mean(r > 0, na.rm = TRUE) * 100, 2)
  )
}

tibble::tibble(
  Metric = c("Total Return (%)", "Annualised Return (%)", "Sharpe Ratio",
             "Omega Ratio", "Sortino Ratio", "Max Drawdown (%)",
             "Calmar Ratio", "Hit Rate (%)"),
  Strategy = bench_metrics(
    test_strat_r,
    INITIAL_CAPITAL,
    dplyr::last(test_results$portfolio_value)
  ),
  SPY = bench_metrics(
    test_spy_r,
    INITIAL_CAPITAL,
    dplyr::last(spy_test_bench$spy_value)
  ),
  `Equal-Weight` = bench_metrics(
    test_ew_r,
    INITIAL_CAPITAL,
    dplyr::last(ew_bench$ew_value)
  ),
  XLK = bench_metrics(
    test_xlk_r,
    INITIAL_CAPITAL,
    dplyr::last(xlk_bench$xlk_value)
  )
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Testing Period Performance: Strategy vs Benchmarks",
    subtitle = "January 2025 to March 2026. Optimized parameters. No refitting."
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#f0f4ff"),
    locations = gt::cells_body(columns = Strategy)
  ) %>%
  gt::tab_source_note("Equal-Weight = daily equal-weight average of all 11 sector ETFs. XLK = Technology ETF only. Sharpe/Sortino annualised using sqrt(252).")
```

The performance table reinforces a mixed but honest verdict. On every risk-adjusted metric the strategy leads: Sharpe of 0.906 versus SPY's 0.499, Omega of 1.187 versus SPY's 1.106, Calmar of 0.982 versus SPY's 0.417, and a maximum drawdown of -5.6% against SPY's -18.8%. These numbers reflect a portfolio that genuinely managed downside risk better than any of the benchmarks shown. However, the strategy also trailed all three benchmarks on total return and annualised return, which is the harder result to dismiss. The Equal-Weight benchmark is a particular concern: it required no signal generation, no clustering, no Kalman filter, and no optimization, yet it returned 14.7% with a reasonable Omega of 1.177 and a Calmar of 0.783. That result raises a direct challenge to the value of complexity in this model. The most honest interpretation is that the strategy achieved its stated goal of capital preservation with favourable risk-adjusted metrics, but in a market environment where equities broadly trended upward over the test period, a lower-beta, more defensive posture was penalised on absolute return. **In the long run, equity markets go up, and a strategy that systematically reduces market exposure will likely underperform passive benchmarks over full cycles unless the risk reduction is severe enough to compound through avoided drawdowns** -- and 15 months is too short a window to know whether that is the case here.

### 10.5 Late-2025 Zoom: October 29, 2025 Onward (Indexed to 100)

```{r test-zoom-chart}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 10
#| fig-height: 5

ZOOM_START <- as.Date("2025-10-29")

combined_test %>%
  dplyr::filter(date >= ZOOM_START) %>%
  dplyr::mutate(
    strat_z = portfolio_value / dplyr::first(portfolio_value) * 100,
    spy_z   = spy_value       / dplyr::first(spy_value)       * 100,
    ew_z    = ew_value        / dplyr::first(ew_value)        * 100,
    xlk_z   = xlk_value       / dplyr::first(xlk_value)       * 100
  ) %>%
  dplyr::select(date,
                Strategy       = strat_z,
                SPY            = spy_z,
                `Equal-Weight` = ew_z,
                XLK            = xlk_z) %>%
  tidyr::pivot_longer(-date, names_to = "series", values_to = "value") %>%
  ggplot(aes(x = date, y = value, colour = series, linewidth = series)) +
  geom_line(alpha = 0.9) +
  scale_colour_manual(values = c(
    "Strategy"     = "#2c7bb6",
    "SPY"          = "#d7191c",
    "Equal-Weight" = "#1a9641",
    "XLK"          = "#f4a736"
  )) +
  scale_linewidth_manual(values = c(
    "Strategy"     = 1.0,
    "SPY"          = 0.6,
    "Equal-Weight" = 0.6,
    "XLK"          = 0.6
  )) +
  labs(
    title    = "Late-2025 Zoom: Strategy vs Benchmarks (Indexed to 100)",
    subtitle = "October 29, 2025 to March 2026. All series re-indexed to 100 at start of window.",
    x        = NULL, y = "Index (100 = Oct 29 2025)", colour = NULL, linewidth = NULL,
    caption  = "Equal-Weight = buy-and-hold equal allocation across all 11 sector ETFs. XLK = Technology ETF."
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

It is worth being transparent about what this chart is doing: October 29, 2025 was chosen because it roughly coincides with where technology peaked before a multi-month decline, which flatters the strategy's relative performance. That is a form of **storytelling bias**: selecting a start date that makes the result look better than the full-period picture. Even acknowledging that bias explicitly, the strategy still did not outperform Equal-Weight over this window. Equal-Weight held its gains better through the late-2025 period and finished above the strategy's index level despite having no defensive mechanism at all. The strategy did protect capital relative to SPY and XLK during the drawdown period, which is consistent with its design, but the honest conclusion is that a simple passive equal allocation across all eleven sector ETFs was a harder benchmark to beat than anticipated. This zoomed window, even when framed as favourably as possible, does not change that verdict.

------------------------------------------------------------------------

## 11. Lessons Learned and Conclusion

### 11.1 Rolling Windows Fix the FIN 440 Fixed-Parameter Problem, But Threshold Selection Remains Hard

In FIN 440, one of the clearest failures was setting fixed parameters over a multi-year window without any rolling re-estimation. A single threshold applied from 2022 to 2024 was inevitably calibrated to the dominant regime of the training period and failed when conditions changed. This project addressed that directly by implementing a 60-day rolling lookback window for cluster return computation and a 20-day SMA on cumulative returns, meaning the signal adapts continuously to recent market behavior rather than being anchored to a static historical level. However, the threshold problem was not fully solved: the signal threshold of 0.050 was still selected as a single fixed value from the optimization surface, and as the testing period demonstrated, a threshold calibrated on 24 years of training data may be too conservative for the tighter spread environment of a specific 15-month window. The lesson is that rolling windows are necessary but not sufficient; the thresholds that gate entry and exit also need to be adaptive, not fixed.

### 11.2 Storytelling Bias Is Subtle and Must Be Named Explicitly

In FIN 440, the observation that "the model performs better in volatile situations" was identified as storytelling bias: a post-hoc narrative built around what the data happened to show, rather than a feature that was intentionally designed. This project made the same mistake in Section 10.5, where the late-2025 zoom window was chosen to start on October 29, 2025, a date that coincided with the technology sector peak and therefore flattered the strategy's relative performance. The difference in this submission is that the bias was named and acknowledged directly in the commentary, rather than presented as evidence of the strategy's strength. Calling out storytelling bias explicitly does not eliminate it, but it is the honest practitioner's responsibility to distinguish between "this is what the data shows" and "this is a feature I designed and can reproduce." Every performance chart should be interrogated with the question: would this chart look the same if I chose a different start date?

### 11.3 Outlier Sensitivity Was Tested This Time: COVID Robustness Check as a Direct Response to FIN 440

In FIN 440, model parameters were heavily influenced by major outliers in the training period. A single dominant event shaped the optimization surface in ways that were not visible until the model was deployed out-of-sample. This project explicitly addressed that failure with a COVID robustness check in Section 9.6, re-running the full threshold optimization with the February to June 2020 period excluded from performance measurement. The result was that the optimal parameters were nearly identical with and without the exclusion window, which provides meaningful evidence that the selected thresholds are not simply "COVID-proof parameters" that exploit one extreme event. The broader lesson is that any training period containing one or two extreme events should be treated with skepticism until the optimizer has been shown to produce consistent results with and without those events.

### 11.4 Signal Activity Must Be Validated: One Trade in 15 Months Is Not a Strategy

The most uncomfortable result in the testing period is that the strategy executed exactly one trade over 15 months. The signal threshold of 0.050 was selected from the training optimization surface, but the realized spread between the Cyclical and Defensive clusters in the testing environment rarely reached that level, a consequence of the relatively low cross-sectional volatility of mid-to-late 2025. This is a failure mode that does not show up clearly in equity curves or performance tables, because a portfolio holding a prior position can still earn returns; the failure is that the strategy is not doing what it was designed to do. Any threshold-gated rotation strategy must be validated not just on risk-adjusted return metrics but on signal fire rate, average holding period, and number of state transitions. In future work, the threshold would be set as a percentile of the rolling spread distribution rather than a fixed level, so the signal frequency adapts to the volatility regime.

### 11.5 Complexity Did Not Beat the Simplest Benchmark

The most humbling result of this project is that a plain equal-weight buy-and-hold allocation across all eleven sector ETFs, requiring no clustering, no Kalman filter, no optimization, and no signal logic, outperformed the strategy on total return, annualized return, and most absolute metrics over the testing period. The strategy's edge was entirely in risk management: lower maximum drawdown (-5.6% vs -15.0% for Equal-Weight), better Sharpe, Omega, and Calmar ratios. In a rising market environment, lower volatility and lower drawdown come at the cost of lower participation in the upside, and over a 15-month test period in a broadly positive equity market, that tradeoff penalized the strategy significantly. This is not a reason to abandon the design, but it is a reason to be honest about what the strategy is: a capital preservation vehicle, not a return maximization engine. The lesson from FIN 440 and FIN 452 together is that model complexity must justify itself against simple baselines, and "it has a better Sharpe" is a weaker argument when the absolute return gap is large.

### 11.6 The Kalman Filter Responsiveness and Signal Threshold Compound Each Other

The Kalman filter was designed to act as an adaptive proxy for a 200-day moving average, serving as the slow-moving long-run baseline against which short-term cluster momentum could be measured. The logic was straightforward: a 20-day SMA captures where the cluster has been recently, while the KF level captures where it has been over a much longer horizon. The spread between them, SMA minus KF level, would be positive when recent momentum is running ahead of the long-run trend and negative when the cluster is underperforming its baseline. This is conceptually identical to a moving average crossover signal, except the KF was chosen over a fixed 200-day SMA because it weights observations optimally given the noise in the series rather than treating every day in the window equally. The appeal of the design is that a single spread metric captures both the direction and the magnitude of short-term divergence from the long-run trend, giving a clean, economically interpretable signal for when to rotate into or out of a cluster.

The Kalman filter in this strategy uses a dW/dV ratio of 0.001, which makes the filtered level an extremely slow-moving baseline. Because the KF level barely moves from day to day, the spread between the SMA and the KF level reflects deviation from a near-permanent long-run trend rather than any shorter-term relative divergence between the two clusters. This is conservative by design, but it has a compounding effect when combined with a signal threshold of 0.050: the spread has to be both large in absolute terms and sustained over multiple days before any signal fires. In a low cross-sectional volatility environment like mid-to-late 2025, neither condition was frequently met, which is a key reason the strategy executed only one trade in 15 months. The fix is not simply to lower the threshold or raise the KF responsiveness in isolation; both parameters need to be co-optimized, because a more responsive KF (say dW/dV = 0.005 or 0.01) produces a noisier spread that requires a recalibrated threshold to avoid excessive false signals. This interaction between the filter and the threshold was not modeled in the optimization grid, which searched over thresholds with KF responsiveness fixed, and including KF_DW_SCALE as a third optimization dimension is the single highest-priority improvement for a future version of this strategy.

### 11.7 Within-Cluster Momentum Was Ignored, and It Cost Returns

The strategy treats all ETFs within a cluster as interchangeable: once the Cyclical cluster is selected, every ETF in that group receives an equal share of the cyclical allocation. This completely ignores the fact that individual sectors within a cluster can diverge dramatically from one another. During the test period, XLK was the clear standout within the Cyclical cluster, driven by the AI investment theme, while other cyclical sectors like XLE and XLB lagged significantly. A strategy that rotates into the Cyclical cluster without distinguishing between its members will always capture the average, never the best. An improvement would be to add a within-cluster momentum layer: rank ETFs inside each cluster by recent return momentum and tilt the allocation toward the top performers rather than splitting equally. This would have meaningfully increased participation in the technology rally without requiring any change to the cluster structure or signal logic, and it is a natural next iteration that preserves the core architecture while adding a second level of return-seeking behaviour.

### 11.8 The 80/20 Cluster Allocation Should Have Been Optimized, Not Fixed

The 80/20 split between the selected and non-selected cluster was chosen as a diversification rule: always keep some exposure to both sides to avoid concentration risk. In practice, this rule costs conviction. When the strategy signals Momentum Cyclical, it puts 80% in the Cyclical cluster and 20% in Defensive, meaning it is always partially hedged against its own call. A trading model that has generated a signal should act on that signal with appropriate sizing, not automatically diversify away 20% of it regardless of signal strength. The 80/20 was never optimized as a parameter, which means it was an assumption masquerading as a design choice. Including the cluster allocation split in the optimization grid, allowing it to range from 60/40 to 100/0 depending on what the training data supports, would have produced a more honest and potentially more profitable position sizing rule. In markets where conviction is rewarded, playing it safe by construction is just a slower way to underperform.

### 11.9 Every Failure Is a Step Forward

A year ago, building a quantitative trading strategy from scratch, with clustering, Kalman filtering, rolling signals, parallel optimization, and an execution loop with dividends and transaction costs, would have been an entirely foreign task. The FIN 440 model failed on fixed parameters, outlier sensitivity, and no significance testing. This model addressed all three of those failures by design and introduced new failures of its own: infrequent trading, a threshold calibrated to the wrong regime, and a complexity premium that was not earned in the test period. That progression, from not knowing what the mistakes are, to making intentional design choices that solve the last round of mistakes, to identifying a new set of more sophisticated problems, is what learning actually looks like. The model did not succeed in beating passive benchmarks, but the understanding of why it failed is worth more than a model that worked by accident.

### 11.10 Conclusion

I have improved drastically over my FIN 440 model, but there is still a ways to go. This is a testament that in order to succeed in anything, you need to try, try, and try again. Building a model is like a free lottery ticket, except the odds of striking gold increase every time as I continue to improve on past mistakes.
