---
title: "Cluster-Based Relative Value Strategy: Intra-Cluster Mean Reversion on SPDR Sector ETFs"
author: "Mathew Frame"
date: "07/04/2026"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
---

```{r setup}
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(tidyquant)
library(gt)
library(cluster)
library(dlm)
library(ggrepel)
library(scales)
library(PerformanceAnalytics)
library(factoextra)
library(ggdendro)
library(patchwork)
library(plotly)
library(foreach)
library(doParallel)
library(RTL)
```

```{r data-pull}
#| message: false
#| warning: false
#| echo: false

tickers <- c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")

raw <- tidyquant::tq_get(tickers, from = "2000-06-01", to = "2026-03-31")

# Log returns grouped by symbol
returns <- raw %>%
  dplyr::group_by(symbol) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(ret = log(adjusted / dplyr::lag(adjusted))) %>%
  dplyr::ungroup() %>%
  dplyr::filter(!is.na(ret))

# Training / testing split -- date filter only, never slice()
TRAIN_END  <- "2024-12-31"
TEST_START <- "2025-01-01"

ret_train <- returns %>% dplyr::filter(date <= TRAIN_END)
ret_test  <- returns %>% dplyr::filter(date >= TEST_START)

# Sector ETFs only (SPY is benchmark, not traded)
sector_tickers <- tickers[tickers != "SPY"]

sector_train <- ret_train %>% dplyr::filter(symbol %in% sector_tickers)
sector_test  <- ret_test  %>% dplyr::filter(symbol %in% sector_tickers)
spy_train    <- ret_train %>% dplyr::filter(symbol == "SPY")
spy_test     <- ret_test  %>% dplyr::filter(symbol == "SPY")
```

------------------------------------------------------------------------

## Executive Summary

*\[To be written last. 3 sentences: strategy thesis, training result, testing result.\]*

------------------------------------------------------------------------

## 1. Intuitive Rationale

### Strategy Thesis

This strategy tests one hypothesis: that sector ETFs temporarily deviate from their co-moving peers and that those deviations are more likely to revert than persist. The mechanism being tested is whether unsupervised clustering on rolling return correlations can identify which ETFs genuinely move together, and whether overweighting the recent underperformer within each group generates risk-adjusted returns above a passive SPY benchmark. This project does not assume the hypothesis is correct. It tests it honestly, with full acknowledgment of where and why it fails.

### What Clustering Adds

A naive mean-reversion strategy would buy any ETF that has dropped recently. The problem is that not all ETFs are comparable: buying XLE because it underperformed XLK assumes those two sectors share a common driver, which they often do not. Clustering first identifies which ETFs are genuinely co-moving in the current market environment, based on 60 days of rolling return correlations. Only within those empirically identified peer groups does the mean-reversion signal have a structural basis. The cluster membership itself is dynamic: it updates monthly as correlations shift, so the strategy adapts to changing market structure rather than assuming fixed relationships.

### Why the Kalman Filter Is Required

Within-cluster underperformance can have two causes: temporary noise (the ETF will revert) or a genuine trend (the ETF is repricing to a new level). Buying into a downtrend on the assumption it will revert is the single largest failure mode of mean-reversion strategies. The Kalman Filter estimates the underlying trend in each ETF's price series by smoothing out daily noise. If the smoothed slope is negative beyond a threshold, the ETF is excluded from the portfolio entirely that day, regardless of how undervalued it appears relative to its cluster. The Kalman Filter is therefore not a signal enhancer; it is a risk gate that prevents the strategy from confusing structural repricing with temporary deviation.

### When This Strategy Fails

This strategy is designed for range-bound, mean-reverting environments and will fail in two specific conditions. The first is sustained directional trends: when sectors are repricing to fundamentally new levels (the 2022 rate shock driving XLU and XLRE down for months), within-cluster underperformers continue underperforming and the mean-reversion bet loses repeatedly. The Kalman gate reduces this exposure but does not eliminate it entirely, particularly at the onset of a trend before the slope signal has confirmed direction. The second is correlation collapse: during systemic shocks (March 2020, September 2008), all 11 ETFs correlate toward 1.0, the cluster structure becomes meaningless, and within-cluster z-scores carry no information. The strategy's cluster quality filter addresses this by disabling trading when average intra-cluster correlation falls below a minimum threshold.

### Academic Foundation

Two papers motivate the design. Ajayi (2025) demonstrates that unsupervised K-Means clustering identifies the same major market regimes as parametric Markov Switching models, validating that rolling clustering captures economically meaningful structural breaks without requiring distributional assumptions. Raju (2025) uses rolling K-Means on equity return data and finds that cluster-based signals generate statistically significant return differentials across regimes (Sharpe differential of 2 to 4 points, p \< 0.01), though those results are from Indian equities where return synchronicity is higher than in U.S. markets. Morck, Yeung, and Yu (2000) show that U.S. stocks have low return synchronicity relative to emerging markets, meaning firm-specific and sector-specific variation dominates macro signals. This implies that intra-cluster dispersion in U.S. sector ETFs will be meaningful and tradeable, but that Raju's Sharpe differentials represent an upper bound rather than a replication target.

------------------------------------------------------------------------

## 2. Data and Features

```{r data-sources-table}
#| message: false
#| warning: false
#| echo: false

tibble::tibble(
  Ticker      = c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
                  "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY"),
  Name        = c("SPDR S&P 500 ETF Trust", "Materials", "Communication Services",
                  "Energy", "Financials", "Industrials", "Technology",
                  "Consumer Staples", "Real Estate", "Health Care",
                  "Utilities", "Consumer Discretionary"),
  Role        = c("Benchmark", rep("Traded", 11)),
  `Available From` = c("2000-06-01", "2000-06-01", "2018-06-19", "2000-06-01",
                        "2000-06-01", "2000-06-01", "2000-06-01", "2000-06-01",
                        "2015-10-08", "2000-06-01", "2000-06-01", "2000-06-01"),
  Source      = rep("Yahoo Finance via tidyquant::tq_get()", 12)
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Universe: SPDR Sector ETFs and Benchmark",
    subtitle = "Daily adjusted close prices. Training: June 2000 to December 2024. Testing: January 2025 to March 2026."
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#f0f4ff"),
    locations = gt::cells_body(rows = Role == "Benchmark")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(columns = Ticker)
  ) %>%
  gt::tab_source_note("XLC available from June 2018; XLRE available from October 2015. Both included despite shorter histories.")
```

All price data is sourced via API with no local files. Log returns are computed as $r_t = \log(P_t / P_{t-1})$ on adjusted close prices, which account for dividends and splits. Log returns are used rather than simple returns because they are additive across time, approximately normally distributed, and produce a more stable correlation matrix, which is the direct input to the clustering algorithm. SPY is carried as the benchmark throughout but is never included in the traded universe or the clustering.

The clustering input is not the raw return series but the pairwise correlation matrix computed over a rolling 60-day window. Clustering directly on returns would group ETFs by recent absolute performance, crowding recent winners into one cluster regardless of whether they share underlying economic drivers. Correlation captures co-movement independent of magnitude: two ETFs are similar if they move in the same direction at the same time, not because they both went up. The 60-day window is a deliberate balance between responsiveness and stability. A shorter window (20 days) produces noisy correlations dominated by single events; a longer window (252 days) is too slow to reflect structural shifts in market relationships and would keep the portfolio anchored to stale cluster assignments. The 60-day parameter is treated as tunable and will be evaluated in the optimization section.

------------------------------------------------------------------------

## 3. Clustering

### 3.1 Cluster Hypothesis

The following cluster assignments are stated before any computation is performed. Results will be assessed honestly against these priors.

```{r cluster-hypothesis}
#| message: false
#| warning: false
#| echo: false

tibble::tibble(
  Cluster       = c("Cyclical", "Defensive", "Commodity", "Rate-Sensitive"),
  Sectors       = c("Industrials, Financials, Consumer Discretionary",
                    "Staples, Utilities, Healthcare",
                    "Energy, Materials",
                    "Technology, Communication Services, Real Estate"),
  `Core Driver` = c("GDP / growth",
                    "Stability / recession",
                    "Inflation / commodities",
                    "Interest rates")
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Cluster Hypothesis: Expected Groupings Before Computation",
    subtitle = "Stated prior to running the elbow method or hierarchical clustering."
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) %>%
  gt::cols_width(Cluster        ~ px(160),
                 Sectors        ~ px(300),
                 `Core Driver`  ~ px(200)) %>%
  gt::tab_source_note("Hypothesis informed by sector economic characteristics, not by training data.")
```

### 3.2 Selecting k: Elbow Method

```{r elbow-method}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 8
#| fig-height: 4

# Compute average correlation matrix over full training period
ret_wide_train <- sector_train %>%
  dplyr::select(date, symbol, ret) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = ret) %>%
  dplyr::arrange(date) %>%
  dplyr::select(-date)

cor_mat  <- cor(ret_wide_train, use = "pairwise.complete.obs")
dist_mat <- as.dist(1 - cor_mat)

# Hierarchical clustering
hc <- hclust(dist_mat, method = "ward.D2")

# WSS for k = 2 to 8
compute_wss <- function(k) {
  labels    <- cutree(hc, k = k)
  centroids <- tapply(seq_along(labels), labels, function(idx) idx)
  total_wss <- sum(sapply(unique(labels), function(cl) {
    members <- names(labels[labels == cl])
    if (length(members) < 2) return(0)
    sub_dist <- as.matrix(dist_mat)[members, members]
    sum(sub_dist^2) / (2 * length(members))
  }))
  tibble::tibble(k = k, wss = total_wss)
}

wss_df <- purrr::map_dfr(1:8, compute_wss)

ggplot(wss_df, aes(x = k, y = wss)) +
  geom_line(colour = "#2c7bb6", linewidth = 0.9) +
  geom_point(colour = "#2c7bb6", size = 3) +
  geom_vline(xintercept = 3, linetype = "dashed", colour = "#d7191c", linewidth = 0.8) +
  annotate("text", x = 3.2, y = max(wss_df$wss) * 0.95,
           label = "Selected k = 3", colour = "#d7191c", hjust = 0, size = 3.5) +
  scale_x_continuous(breaks = 1:8) +
  labs(
    title    = "Elbow Method: Within-Cluster Sum of Squares vs. Number of Clusters",
    subtitle = "Training period average correlation matrix. Ward.D2 hierarchical clustering.",
    x        = "Number of Clusters (k)",
    y        = "Within-Cluster Sum of Squares (WSS)",
    caption  = "Distance metric: d(i,j) = 1 - corr(i,j). Training period June 2000 to December 2024."
  ) +
  theme_minimal(base_size = 11)
```

Despite my initial hypothesis being k = 4, **the elbow I see is at k = 3.** There is a meaningful reduction in WSS from 3 -\> 4, but the decision to use k = 3 comes from the change in slope from 2 -\> 4 having a sharper angle than 3 -\> 5. There is no right answer here, but using 4 clusters creates the risk of weak clusters that change often, leading to increased transaction costs for the trading strategy.

### 3.3 Hierarchical Clustering

*\[To be built after k is confirmed from elbow method above.\]*

------------------------------------------------------------------------

## 4. Kalman Filter

*\[Local level model, smoothed price and slope, hard gate rule, visualization, gate firing frequency.\]*

------------------------------------------------------------------------

## 5. Signal Generation

*\[Within-cluster z-scores, inverse rank weighting, cross-cluster equal weight, worked example.\]*

------------------------------------------------------------------------

## 6. Execution and Implementation

*\[Mermaid mental model, execution rules, signal lag, transaction costs, position sizing.\]*

------------------------------------------------------------------------

## 7. Training Period Backtest

*\[Equity curve vs SPY, drawdown chart, performance table.\]*

------------------------------------------------------------------------

## 8. Risk Appetite

*\[Max drawdown tolerance, optimization objective, defensive trigger.\]*

------------------------------------------------------------------------

## 9. Optimization

*\[Grid search over parameters, heat map, selected parameters.\]*

------------------------------------------------------------------------

## 10. Testing Period Results

*\[Out-of-sample backtest, honest assessment.\]*

------------------------------------------------------------------------

## 11. Lessons Learned

*\[Min 5 points x 3-5 sentences, FIN 440 QTP failures referenced.\]*
